---
title: "axial_SphereOfInfluence_code"
format: html
editor: visual
---

Moving from just the plume profile and adding in more samples to instead look at "sphere of influence".

#Setting up environment

```{r}
#Expanded sample list from plume profile to sphere of influence

library(readr)
library(tidyverse)
library(phyloseq)
library(readxl)

#Import data
all_metadata_jan2026 <- read_csv("~/R/Axial2023/datainput/all_metadata_jan2026_v6.csv")
axial_decontam_DF <- read_excel("~/R/Axial2023/datainput/axial_decontam_DF.xlsx")
ASV <- read.delim("C:/Users/alexis98adams/Documents/R/Axial2023/datainput/clean_ASV.tsv",
                  header = TRUE, sep = "\t")
tax <- read.delim("C:/Users/alexis98adams/Documents/R/Axial2023/datainput/taxonomy.tsv", header = TRUE, sep = "\t")

samplestokeep_updated <- read_csv("~/R/Axial2023/datainput/axial_samplestokeep_updated.csv")

axial_euk_cell_counts <- read_csv("~/R/Axial2023/datainput/axial_euk_cell_counts.csv")

#Make metadata sheet and fix naming
tax <- tax %>% rename(FeatureID = Feature.ID)

all_metadata_jan2026 <- all_metadata_jan2026 %>% 
  rename(Sample = X)

axial_metadata <- all_metadata_jan2026 %>% filter(site == "Axial") %>% 
  filter(Sample %in% samplestokeep_updated$Sample) 

###Adjust sample name and identify replicates
axial_metadata_wReps <- axial_metadata %>% 
  mutate(DepthCat_Location = paste(dive_ctd, filter, sample_name, location_name, depth_category, depth, sep = "-"))

#write.csv(axial_metadata_wReps, "axial_metadata_wReps.csv", row.names = FALSE)

axial_metadata_wReps <- read_csv("~/R/Axial2023/axial_metadata_wReps.csv")

#Create two different sample identifiers for grouping (one general, one specific)
axial_metadata_wReps <- axial_metadata_wReps %>% 
  mutate(sample_type = case_when(
    depth_category == "above vent" ~ "Above Vent", 
    depth_category == "oxycline" ~ "Water Column", 
    depth_category == "euphotic" ~ "Water Column",
    depth_category == "diffuse" ~ "Diffuse Vent", 
    depth_category == "30m above" ~ "Water Column", 
    str_detect(depth_category, "plume") ~ "Water Column",
    depth_category == "background" ~ "Background",
    TRUE ~ depth_category)) %>%
  mutate(sample_name = case_when(
    sample_name == "SoupyBSW" ~ "Caldera Basin Water", 
    TRUE ~ sample_name)) %>% 
  mutate(depth_location = paste(sample_type, sample_name, depth, sep = "-")) %>% 
  mutate(depth_location = case_when(
    str_detect(depth_location, "Background-Background-1200") ~ "Background-1200", 
    str_detect(depth_location, "Background-Background-1300") ~ "Background-1300",
    str_detect(depth_location, "Background-Background-1400") ~ "Background-1400",
    TRUE ~ depth_location)) %>% 
  mutate(depth_location_wReps = paste(depth_location, Rep, sep = "-"))

```

#Env heatmap

```{r}
#Don't use the table with ASV data because you will loose high temp vents

env_data <- axial_metadata_wReps %>%
  select(depth_location, depth, eukCONC, prok, pH, TOC, TDN,
         Cl, SO4, Na, K, Mg, `18S`, Ca, temp) %>%
  distinct() %>% 
  mutate(`18S` = as.character(`18S`)) %>% 
  mutate(
    depth_location = reorder(depth_location, -depth)) #order by depth_location using the depth column

variable_fill <- list(
  "Depth" = "depth",
  "Temp" = "temp",
  "Euk Cell Count" = "eukCONC",
  "Prok Cell Abundance" = "prok",
  "pH" = "pH",
  "TOC" = "TOC",
  "TDN" = "TDN",
  "Cl" = "Cl",
  "SO4" = "SO4",
  "Na" = "Na",
  "K" = "K",
  "Mg" = "Mg",
  "Ca" = "Ca",
  "18S" = "18S")

#Set scales
scale_fill <- list( 
  scale_fill_gradient(low = "#ADD", high = "#191970"), # depth
  scale_fill_gradient(low = "lightblue", high = "coral"), #temp
  scale_fill_gradient(low = "#d8d8f0", high = "#DA70D6"), # eukCONC
  scale_fill_gradient(low = "#d8d8f0", high = "#8B008B"), # prok
  scale_fill_gradient(low = "#B22", high = "#8A2BE2"), # pH
  scale_fill_gradient(low = "#efa", high = "#FF7F50"), # TOC
  scale_fill_gradient(low = "#e38aa6", high = "#D77"), # TDN
  scale_fill_gradient(low = "#DAA", high = "#ba4e70"), # Cl
  scale_fill_gradient(low = "#a13356", high = "#8a1a3d"), # SO4
  scale_fill_gradient(low = "#EEE", high = "#BDB"), # Na
  scale_fill_gradient(low = "#EEE8AA", high = "#BDB76B"), # K
  scale_fill_gradient(low = "#7B68EE", high = "#483D8B"), # Mg
  scale_fill_gradient(low = "indianred", high = "orange"), #Ca
  scale_fill_manual(values = c("0" = "#F5FFFA", "1" = "#000000"))) #18S

names(scale_fill) <- names(variable_fill)

env_heatmap <- ggplot(env_data, aes(y = depth_location)) +
  purrr::imap(variable_fill, \(x, y) {
    list(
      geom_tile(
        aes(x = y, fill = .data[[x]]),
        color = "black",
        width = 0.95),
      scale_fill[[y]],
      ggnewscale::new_scale_fill())}) +
  theme_void() +
  theme(
    axis.text.y = element_text(hjust = 1),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_blank(),
    legend.position = "right",
  legend.box = "vertical",
  legend.direction = "horizontal")

ggsave("~/env_heatmap.png", plot = env_heatmap, width = 10, height = 10, dpi = 600, units = "in")
```

#Stacked bar plot of relative abundance

```{r}
#Keep ASVs that only have 50 or more sequences and are found in two or more samples
asv_keep <- ASV %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "sample",
    values_to = "SEQUENCE_COUNT") %>%
  group_by(FeatureID) %>%
  summarise(
    total_reads = sum(SEQUENCE_COUNT),
    prevalence = sum(SEQUENCE_COUNT > 0),
    .groups = "drop") %>%
  filter(total_reads >= 50, prevalence >= 2)

samplestokeep_ASV <- ASV %>% 
  filter(FeatureID %in% asv_keep$FeatureID) %>% 
  pivot_longer(
    cols = -c(FeatureID),
    names_to = "Sample",
    values_to = "SEQUENCE_COUNT") %>%
  filter(Sample %in% axial_metadata$Sample) 

table_asv <-  samplestokeep_ASV %>% 
  left_join(axial_metadata, by = "Sample") %>% 
  left_join(tax, by = "FeatureID") %>% 
  separate(
    col = Taxon,
    into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Genus", "Species"),
    sep = ";")

alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")

#Curate taxonomy
#library(dplyr)
#library(stringr)
#library(purrr)

tax_asv_metadata <- table_asv %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(Phylum != "Opisthokonta") %>% 
  mutate(
    Phylum = case_when(
      str_detect(Phylum, "_X$") ~ "Unknown",
      is.na(Phylum) ~ "Unknown",
      TRUE ~ Phylum),
    Class = case_when(
      str_detect(Class, "_X$") ~ "Unknown",
      is.na(Class) ~ "Unknown",
      TRUE ~ Class),
    Genus = case_when(
      str_detect(Genus, "_X") ~ "Unknown",
      is.na(Genus) ~ "Unknown",
      TRUE ~ Genus),
    Order = case_when(
      str_detect(Order, "_X") ~ "Unknown",
      is.na(Order) ~ "Unknown",
      TRUE ~ Order)) %>% 
  rowwise() %>% #Taxonomy to the order level
  mutate(
    taxonomy_collapse = paste(
      na.omit(c(
        if (Phylum != "Unknown") Phylum, 
        if (Class != "Unknown") Class,
        if (Genus != "Unknown") Genus, 
        if (Order != "Unknown") Order)),
      collapse = ";")) %>% 
   mutate(phylum_class = paste(Phylum, Class, sep = "-")) %>% 
  mutate(taxa = case_when( #Curate taxonomy
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Class == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Class == "Radiolaria" ~ "Rhizaria-Radiolaria",
    phylum_class == "Unknown-Unknown" ~ "Unknown Eukaryota",
    str_detect(phylum_class, "Unknown-Amoebozoa_XX|Tubulinea|Evosea") ~ "Amoebozoa",
    phylum_class %in% alv ~ "Alveolata-Other",
    phylum_class == "Stramenopiles-Unknown" ~ "Stramenopiles-Other",
    str_detect(phylum_class, "Discoba|Metamonada") ~"Excavata",
    str_detect(phylum_class, "Cryptophyta|Kathablepharida") ~"Cryptista",
    str_detect(phylum_class, "Chlorophyta|Prasinodermophyta|Picozoa") ~"Archaeplastida",
    str_detect(phylum_class, "Apusomonada|Breviatea|Amoebozoa") ~"Amorphea",
    str_detect(phylum_class, "Centroplasthelida|Haptophyta") ~"Haptista",
    str_detect(phylum_class, "Nibbleridia") ~"Provora-Nibbleridia",
    str_detect(phylum_class, "Collodictyonidae|Mantamonadidea") ~ "CRuMS",
    str_detect(phylum_class, "Telonemia") ~ "Telonemia",
    TRUE ~ phylum_class)) %>% 
  left_join(
  axial_metadata_wReps %>% 
    select(Sample, depth_location, depth_location_wReps),
  by = "Sample") 

#If plotting using broader depth_location this collapses replicates so sum sequences.
#Aggregate ASVs into taxonomy groups (taxonomy_collapse) and calculate rel seq abun. 

collapsed_reps_ASV <- tax_asv_metadata %>% 
  group_by(taxonomy_collapse, depth_location) %>% 
  summarise(
    seq_sum = sum(SEQUENCE_COUNT, na.rm = TRUE),
    across(
      c(taxa, phylum_class,
        location_name, sample_name, depth),
      first),
    .groups = "drop") %>% 
  group_by(depth_location) %>% 
   mutate(sample_seq_total = sum(seq_sum),
         taxa_relAbun = seq_sum / sample_seq_total) %>% 
  ungroup() %>% mutate(
    depth_location = reorder(depth_location, -depth))
  

taxa <- c("Alveolata-Dinoflagellata" = "cornflowerblue","Rhizaria-Radiolaria" = 
            "coral4","Alveolata-Ciliophora" = "coral", 
          "Stramenopiles-Gyrista" = "chocolate","Rhizaria-Cercozoa" = "cadetblue","Stramenopiles-Bigyra" = "aquamarine","Archaeplastida" = "firebrick4", 
          "Discosea-Unknown" = "plum4","Stramenopiles-Other" = "lightsalmon", "Amorphea" = "darkolivegreen3","Haptista" = "gold","Amoebozoa" = "lightgoldenrod",
          "Unknown Eukaryota" = "mistyrose","Telonemia" = "rosybrown","Alveolata-Other" = "peachpuff4","Cryptista" = "coral3","CRuMS" = "darkcyan",
          "Excavata" = "orchid","Provora-Nibbleridia" = "powderblue")

bydepth_stackedbar <- ggplot(collapsed_reps_ASV, aes(y = depth_location, x = taxa_relAbun, fill = taxa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = taxa) +
  theme_minimal(base_size = 12) +
  theme(
    # Text
    axis.text.x = element_text(face = "bold", color = "black", size = 12),
    axis.text.y = element_text(face = "bold", color = "black", size = 11),
    axis.title = element_text(color = "black", face = "bold"),
    strip.text = element_text(color = "black", face = "bold", size = 12),
    legend.text = element_text(color = "black", size = 14),
    
    # Backgrounds
     plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    strip.background = element_rect(fill = "gray", color = NA),
    
    # Gridlines
    panel.grid = element_blank(),
    
    # Legend
    legend.position = "right",
    legend.title = element_blank()) +
  labs(
    title = "",
    x = "",
    y = "") +
  guides(fill = guide_legend(title = NULL))

ggsave("~/bydepth_stackedbar.png", plot = bydepth_stackedbar, width = 8, height = 8, dpi = 600, units = "in")

```

#K-means clustering by sequence data. Normalize raw sequence data around samples.

```{r}
#install.packages("anomalize")
#library(anomalize)

#install.packages("cluster")
#library(cluster)

#install.packages("factoextra")
#library(factoextra)

#library(vegan)

set.seed(1776)

#Group by depth_category, location_name, depth. Use depth_location. 
unique(tax_asv_metadata$depth_location)

seq_only_mat <- tax_asv_metadata %>% 
  group_by(depth_location, FeatureID) %>% 
  summarise(
    mean_rep_seq = mean(SEQUENCE_COUNT, na.rm = TRUE),
    .groups = "drop") %>% 
  pivot_wider(names_from = FeatureID, values_from = mean_rep_seq, values_fill = 0) %>%   distinct() %>% 
  drop_na() %>%
  column_to_rownames(var = "depth_location") %>% 
  as.matrix

#Normalize matrix of numeric metadata
# Using clr approach, margin is set to 2 to perform this across each ASV ID, so the relationships between samples can be determined. 
standarize_seq <- decostand(seq_only_mat, MARGIN = 2, method = "rclr")

# Output from here should be stn_niskin
rownames(standarize_seq) <- rownames(seq_only_mat)
colnames(standarize_seq) <- colnames(seq_only_mat)

#Group to see how many clusters are needed
kclusts <-
  tibble(k = 2:10) %>%
  mutate(
    kclust = map(k, ~ kmeans(standarize_seq, .x)),
    glanced = map(kclust, broom::glance))

# Unnest
output_kclusts <- kclusts %>%
  unnest(cols = c(glanced))
# ?kmeans

# plot shows hypothetical number of k clusters (x-axis) vs the within-cluster variance (y-axis). AS we increase the number of clusters, the variance within each cluster get smaller because clustering groups become more significant. 
output_kclusts %>%
  ggplot(aes(x = k, y = tot.withinss)) +
    geom_line(alpha = 0.5, linewidth = 1.2, color = "darkgreen") +
    geom_point(size = 2, color = "darkgreen") +
  theme_linedraw()

#Perform Kmeans with 7 centers
kmeans_seq_7 <- kmeans(standarize_seq, centers = 7)
summary(kmeans_seq_7)

# Report some findings using "broom"
broom::tidy(kmeans_seq_7)

broom::glance(kmeans_seq_7)

# Combine with data
output_kmeans_seq7 <- broom::augment(kmeans_seq_7, standarize_seq)
head(data.frame(output_kmeans_seq7)) 

kmeans_membership_7 <- data.frame(output_kmeans_seq7) %>% 
  select(depth_location = `.rownames`, CLUSTER = `.cluster`) %>% distinct()

# Samples in each cluster
summary(kmeans_membership_7$CLUSTER)

#Create new metadata sheet with cluster info and averaged sequences
depth_order <- tax_asv_metadata %>%
  distinct(depth_location, depth) %>%
  arrange(desc(depth)) %>%
  pull(depth_location)


k_cluster <- tax_asv_metadata %>% 
  select(depth_location, depth, location_name) %>% 
  distinct()  %>% 
  left_join(kmeans_membership_7, by = "depth_location") %>% 
   mutate(CLUSTER = factor(CLUSTER, levels = as.character(1:7))) %>% 
  mutate(depth_location = factor(depth_location, levels = depth_order))

#Visualize which samples are in which clusters

ggplot(k_cluster, aes(x = CLUSTER, y = depth_location)) +
  geom_point(aes(color = location_name), size = 3) +
  scale_color_manual(name = "Vent Field", values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod",
    "Soupy Background" = "lightgoldenrod3",
    "Marker 33" = "mediumpurple",
    "Marker 113" = "mistyrose3")) +
  theme(
    axis.text.y = element_text(hjust = 1),
    axis.text.x = element_text(colour = "black", size = 10), 
    axis.title.x = element_text(size = 11, colour = "black"),
    axis.title.y = element_blank(),
    panel.background = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
    legend.text = element_text(size = 10, colour ="black"), 
    legend.position = "right",
    legend.key.size = unit(0.5, "lines"),
    legend.spacing.y = unit(0.1, "lines"),
    legend.spacing.x = unit(0.1, "lines")) +
  labs(x = "Cluster") 
```


#NMDS
```{r}
#Collapse replicates

asv_wide <- tax_asv_metadata %>% 
  select(FeatureID, SEQUENCE_COUNT, depth_location_wReps) %>% 
  group_by(depth_location_wReps, FeatureID) %>% 
  summarise(seq_avg = mean(SEQUENCE_COUNT), .groups = "drop") %>% 
  pivot_wider(names_from = FeatureID,
              values_from = seq_avg,
              values_fill = 0) %>% 
  column_to_rownames("depth_location_wReps") %>% 
  as.matrix() 

#Compute Bray-Curtis and run NMDS
set.seed(123)
axial_nmds <- metaMDS(asv_wide, distance = "bray") ##stress 0.13 

#Extract NMDS scores
# Extract NMDS coordinates for samples
nmds_scores <- as.data.frame(scores(axial_nmds, display = "sites"))

# Add Sample names as a column
nmds_scores <- nmds_scores %>%
  rownames_to_column("depth_location_wReps")

NMDS_metadata <- tax_asv_metadata %>% 
  left_join(nmds_scores, by = "depth_location_wReps") %>% 
  mutate(sample_name = case_when(
    sample_name == "SoupyBSW" ~ "Caldera Basin Water", 
    sample_name == "IntlDist" ~ "International District",
    TRUE ~ sample_name)) %>% 
  mutate(sample_type = case_when(
    depth_category == "above vent" ~ "Above Vent", 
    depth_category == "oxycline" ~ "Oxycline", 
    depth_category == "euphotic" ~ "Euphotic",
    depth_category == "diffuse" ~ "Diffuse Vent", 
    depth_category == "30m above" ~ "Water Column", 
    str_detect(depth_category, "plume") ~ "Water Column",
    str_detect(location_name, "Soupy") ~ "Caldera Basin Water",
    depth_category == "background" ~ "Background",
    TRUE ~ depth_category))

NMDS <- ggplot(NMDS_metadata, aes(x = NMDS1, y = NMDS2)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(shape = sample_type, color = sample_name), size = 3) +
  scale_color_manual(name = "Sample Location", values = c(
    "Anemone" = "lightgoldenrod",
    "Gollum" = "orange",
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Diva" = "darkcyan",
    "Background" = "darkblue",
    "Caldera Basin Water" = "darkblue",
    "Marker 33" = "mediumpurple",
    "Marker 113" = "mistyrose3")) +
   scale_shape_manual(name = "Sample Location", values = c(
    "Above Vent" = 25,
    "Diffuse Vent" = 24,
    "Water Column" = 19,
    "Oxycline" = 1,
    "Background" = 15,
    "Euphotic" = 1, 
    "Caldera Basin Water" = 0)) +
  theme(
    axis.text.y = element_text(colour = "black", size = 10),
    axis.text.x = element_text(colour = "black", size = 10), 
    axis.title.y = element_text(face = "bold", size = 11), 
    axis.title.x = element_text(face = "bold", size = 11, colour = "black"), 
    panel.background = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
    legend.title = element_text(size = 12, colour = "black", face = "bold"),
    legend.text = element_text(size = 10, colour ="black"), 
    legend.position = "right",
    legend.key = element_blank(),
    legend.key.size = unit(0.5, "lines"),
    legend.spacing.y = unit(0.1, "lines"),
    legend.spacing.x = unit(0.1, "lines")) +
  labs(x = "NMDS1", y = "NMDS2") #+
geom_text(aes(label = depth_location)) 

ggsave("~/NMDS.png", plot = NMDS, width = 6, height = 6, dpi = 600, units = "in")

```


#RDA Analysis
```{r}
set.seed(1998)

#Make a categorical data table 
categorical <- tax_asv_metadata %>% 
  select(depth_location, depth_category, sample_name, sample_type, location_name) %>% 
  distinct() %>% 
    arrange(depth_location) %>% 
  mutate(depth_category = case_when(
    sample_name == "SoupyBSW" ~ "Caldera Bottom Water",
    TRUE ~ depth_category))

#Generate a matrix of numerical data
metadata_RDA <- tax_asv_metadata %>%
  group_by(depth_location) %>%
  summarise(
    depth = first(depth),
    prok = mean(prok),
    temp = mean(temp),
    .groups = "drop"
  ) %>%
  arrange(depth_location) %>%
  column_to_rownames("depth_location") %>% 
  filter(!is.na(prok))


#Normalize matrix of numeric metadata then re-combine with categorical metadata. 
#Standardize compares column by column. Like location_name vs sample type.  Margin is 2. This z-scpres each environmental variable (mean = 0, SD = 1)

norm_metadata <- decostand(metadata_RDA, MARGIN = 2, method = "standardize")

# Combine normalized metadata with categorical data

metadata_standardize <- categorical %>% 
  right_join(as.data.frame(standarize_metadata) %>% rownames_to_column(var = "depth_location")) %>%
  arrange(depth_location) %>% 
  column_to_rownames(var = "depth_location")

```


#Distance decay plot - Jaccard. Does ecological dissimilarity increase with physical separation from diffuse vent?
```{r}
#Calculate both horizontal and vertical distance from source (diffuse)
dist_coords_sample <- read_csv("~/R/Axial2023/dist_coords_sample.csv")

#install.packages("geosphere")
#library(geosphere)

dist_df <- dist_coords_sample %>%
  rowwise() %>%
  mutate(
    horizontal_distance = distHaversine(
      c(long_Sample1, lat_Sample1),
      c(long_Sample2, lat_Sample2))) %>%
  ungroup()

#Calculate total 3D space
dist_df <- dist_df %>%
  mutate(
    distance_3d = sqrt(
      horizontal_distance^2 +
      z_dist^2))

#Format ASV and tax table
asv_matrix <- asv_wide %>% 
  as.data.frame() %>% 
  rownames_to_column("depth_location_wReps") %>% 
  pivot_longer(cols = -depth_location_wReps,
               names_to = "FeatureID",
               values_to = "mean_seq") %>% 
  separate(depth_location_wReps,
    into = c("location","site","depth", "rep"),  
    sep = "-") %>% select(-rep) %>% 
  mutate(SampleID = paste(location, site, depth, sep = "-")) %>% 
  mutate(SampleID = case_when(
  str_detect(SampleID, "Background-1200") ~ "Background-1200",
   str_detect(SampleID, "Background-1300") ~ "Background-1300",
   str_detect(SampleID, "Background-1400") ~ "Background-1400",
    TRUE ~ SampleID)) %>% 
  select(-location, -site, -depth) %>% 
  group_by(FeatureID, SampleID) %>%  
  summarise(seq_sum = sum(mean_seq)) %>% 
  pivot_wider(names_from = SampleID, values_from = seq_sum, values_fill = 0) %>% 
  column_to_rownames("FeatureID") %>% 
  as.matrix()

tax_matrix <- tax %>% 
  separate(Taxon, into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>% select(-Consensus) %>% column_to_rownames("FeatureID") %>% 
  as.matrix()

tax_table <- tax_table(tax_matrix)
ASV_table <- otu_table(asv_matrix, taxa_are_rows = TRUE)

physeq_object <- merge_phyloseq(ASV_table, tax_table, axial_metadata_wReps)

dist <- phyloseq::distance(physeq_object, method = "jaccard")

rownames(as.data.frame(as.matrix(dist)))

dist <- as.data.frame(as.matrix(dist))

dist_scores_df <- dist %>% 
  rownames_to_column(var = "SampleA") %>% 
  pivot_longer(cols = -SampleA, 
               names_to = "SampleB") %>% 
  mutate(Samples_combined = paste(SampleB, SampleA, sep = "-"))

dist_df <- dist_df %>% 
  mutate(Samples_combined = paste(Sample1, Sample2, sep = "-")) %>% 
  left_join(dist_scores_df, by = "Samples_combined")

#Didn't work that well. Manually download and fix. 
write.csv(dist_df, "dist_df.csv", row.names = FALSE)

dist_df_adjusted <- read_csv("~/R/Axial2023/dist_df.csv")

dist_df_refined <- dist_df_adjusted %>% 
  select(SampleA, SampleB, value, distance_3d, Samples_combined)

jaccard_distance <- ggplot(dist_df_refined, aes(x = distance_3d, y = value)) +
  geom_jitter() +
  theme_linedraw() +
  scale_y_continuous(limits = c(0.7,1)) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black", face = "bold", size = 12),
        axis.text = element_text(color = "black", face = "bold", size = 12),
        axis.title = element_text(color = "black", face = "bold", size = 12)) +
  labs(y = "Jaccard Dissimilarity", x = "Euclidean Distance (m)") #+
  geom_text(aes(label = Samples_combined))
  
ggsave("~/jaccard_distance.png", plot = jaccard_distance, width = 6, height = 6, dpi = 600, units = "in")
```

#Jaccard values to dendogram
```{r}
#install.packages("ggdendro")
#library(ggdendro)

dendro_plot_df <- ggdendro::dendro_data(as.dendrogram(cluster_jacc), type = "rectangle")
label_dendro_order <- as.character(dendro_plot_df$labels$label)

dist_plot <- phyloseq::distance(
  physeq_object,
  method = "jaccard",
  binary = TRUE)

hc <- hclust(dist_plot, method = "average")

plot(hc)

cluster_jacc <- hclust(dist_plot, method = "average")

dendro_plot_df <- ggdendro::dendro_data(as.dendrogram(cluster_jacc), type = "rectangle")
label_dendro_order <- as.character(dendro_plot_df$labels$label)
# label_dendro_order

ggplot(segment(dendro_plot_df)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  coord_flip() + 
  scale_y_reverse(expand = c(0.2, 0.5), breaks = c(0, 0.2, 0.4, 0.6, 0.8)) + 
  geom_text(aes(x = x, y = y, label = label, angle = 0, hjust = 0), data = label(dendro_plot_df)) + 
  theme_dendro() + 
  labs(y = "Dissimilarity", title = "Jaccard distance") + 
  theme(axis.text.x = element_text(color = "black", size = 14), 
        axis.line.x = element_line(color = "#252525"), 
        axis.ticks.x = element_line(), axis.title.x = element_text(color = "black", size = 14)) #+ geom_text(aes(label = depth_location))


dendro_plot_output

ggsave("~/ggdendro.png", plot = ggdendro, width = 10, height = 6, dpi = 600, units = "in")

```

