---
title: "axial_allcode_6.29.26"
format: html
editor: visual
---
#Set up environment
```{r}

library(readr)
library(tidyverse)
library(vegan)

all_metadata <- read_csv("~/R/Axial2023/datainput/all_metadata_July2026.csv")

ASV <- read.delim("C:/Users/alexis98adams/Documents/R/Axial2023/datainput/clean_ASV.tsv", header = TRUE, sep = "\t")

tax <- read.delim("C:/Users/alexis98adams/Documents/R/Axial2023/datainput/taxonomy.tsv", header = TRUE, sep = "\t")

samplestokeep_updated <- read_csv("~/R/Axial2023/datainput/axial_samplestokeep_updated.csv")

#axial_euk_cell_counts <- read_csv("R/Axial2023/datainput/axial_euk_cell_counts.csv")

#Make metadata sheet and fix naming
tax <- tax %>% rename(FeatureID = Feature.ID)

all_metadata <- all_metadata %>% 
  rename(Sample = X)

axial_metadata_allsamples <- all_metadata %>% filter(site == "Axial") %>% 
  filter(Sample %in% samplestokeep_updated$Sample) 

#Create two different sample identifiers for grouping (one general, one specific)
axial_metadata_allsamples <- axial_metadata_allsamples %>% 
   mutate(sample_name = case_when(
    sample_name == "SoupyBSW" ~ "Caldera Basin Water", 
    TRUE ~ sample_name)) %>% 
  mutate(sample_type = case_when(
    sample_name == "Caldera Basin Water" ~ "Caldera Basin Water",
    depth_category == "above vent" ~ "Above Vent", 
    depth_category == "oxycline" ~ "Oxycline", 
    depth_category == "euphotic" ~ "Euphotic",
    depth_category == "diffuse" ~ "Diffuse", 
    depth_category == "30m above" ~ "Water Column", 
    str_detect(depth_category, "plume") ~ "Water Column",
    depth_category == "background" ~ "Background",
    TRUE ~ depth_category)) %>% 
  mutate(depth_location = paste(sample_type, sample_name, depth, sep = "-")) %>% 
  mutate(depth_location = case_when(
    str_detect(depth_location, "Background-Background-1200") ~ "Background-1200", 
    str_detect(depth_location, "Background-Background-1300") ~ "Background-1300",
    str_detect(depth_location, "Background-Background-1400") ~ "Background-1400",
    str_detect(depth_location, "Caldera") ~ "Caldera Basin Water - 1535",
    TRUE ~ depth_location)) %>% 
  mutate(location_name = case_when(
    str_detect(depth_location, "Caldera") ~ "Caldera Basin Water",
    TRUE ~ location_name))

axial_metadata_nosurface <- axial_metadata_allsamples %>% 
  filter(!depth_category %in% c("euphotic", "oxycline"))

#Remove ASVs that have less than 100 sequences total and were in two or more samples
#ASV table has 26911 FeatureIDs
#ASV table has 4240 FeatureIDs
asv_morethan50 <- ASV %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "sample",
    values_to = "SEQUENCE_COUNT") %>%
  group_by(FeatureID) %>%
  summarise(
    total_reads = sum(SEQUENCE_COUNT),
    prevalence = sum(SEQUENCE_COUNT > 0),
    .groups = "drop") %>%
  filter(total_reads >= 50, prevalence >= 2)

samplestokeep_ASV <- ASV %>% 
  filter(FeatureID %in% asv_morethan50$FeatureID) %>% 
  pivot_longer(
    cols = -c(FeatureID),
    names_to = "Sample",
    values_to = "SEQUENCE_COUNT") %>%
  filter(Sample %in% axial_metadata_nosurface$Sample) 

table_asv <- samplestokeep_ASV %>% 
  left_join(axial_metadata_nosurface, by = "Sample") %>% 
  left_join(tax, by = "FeatureID") %>% 
  separate(
    col = Taxon,
    into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Genus", "Species"),
    sep = ";") #Still 4240 ASVs


alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")

tax_asv_metadata <- table_asv %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(Phylum != "Opisthokonta") %>% 
  mutate(
    Phylum = case_when(
      str_detect(Phylum, "_X$") ~ "Unknown",
      is.na(Phylum) ~ "Unknown",
      TRUE ~ Phylum),
    Class = case_when(
      str_detect(Class, "_X$") ~ "Unknown",
      is.na(Class) ~ "Unknown",
      TRUE ~ Class),
    Genus = case_when(
      str_detect(Genus, "_X") ~ "Unknown",
      is.na(Genus) ~ "Unknown",
      TRUE ~ Genus),
    Order = case_when(
      str_detect(Order, "_X") ~ "Unknown",
      is.na(Order) ~ "Unknown",
      TRUE ~ Order)) %>% 
  rowwise() %>% #Taxonomy to the order level
  mutate(
    taxonomy_collapse = paste(
      na.omit(c(
        if (Phylum != "Unknown") Phylum, 
        if (Class != "Unknown") Class,
        if (Genus != "Unknown") Genus, 
        if (Order != "Unknown") Order)),
      collapse = ";")) %>% 
   mutate(phylum_class = paste(Phylum, Class, sep = "-")) %>% 
  mutate(taxa = case_when( #Curate taxonomy
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Class == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Class == "Radiolaria" ~ "Rhizaria-Radiolaria",
    phylum_class == "Unknown-Unknown" ~ "Unknown Eukaryota",
    str_detect(phylum_class, "Unknown-Amoebozoa_XX|Tubulinea|Evosea") ~ "Amoebozoa",
    phylum_class %in% alv ~ "Alveolata-Other",
    phylum_class == "Stramenopiles-Unknown" ~ "Stramenopiles-Other",
    str_detect(phylum_class, "Discoba|Metamonada") ~"Excavata",
    str_detect(phylum_class, "Cryptophyta|Kathablepharida") ~"Cryptista",
    str_detect(phylum_class, "Chlorophyta|Prasinodermophyta|Picozoa") ~"Archaeplastida",
    str_detect(phylum_class, "Apusomonada|Breviatea|Amoebozoa") ~"Amorphea",
    str_detect(phylum_class, "Centroplasthelida|Haptophyta") ~"Haptista",
    str_detect(phylum_class, "Nibbleridia") ~"Provora-Nibbleridia",
    str_detect(phylum_class, "Collodictyonidae|Mantamonadidea") ~ "CRuMS",
    str_detect(phylum_class, "Telonemia") ~ "Telonemia",
    TRUE ~ phylum_class))

n_tax_asv_metadata <- tax_asv_metadata %>%
  pull(FeatureID) %>%
  n_distinct() #2911 ASVs

tax_asv_metadata <- tax_asv_metadata %>% 
  mutate(depth_location = paste(sample_type, sample_name, depth, sep = "-"))

#write.csv(tax_asv_metadata, file = "tax_asv_metadata.csv", row.names = FALSE)
```

#Env Heatmap, cell counts and TOC plot
```{r}

#Don't use the table with ASV data because you will loose high temp vents

env_data <- axial_metadata %>%
  select(depth_location, depth, eukCONC, prok, pH, TOC, TDN,
         SO4, `18S`, temp, Cl, Na, K, Mg, Ca) %>%
  distinct() %>% 
  rename(`Depth(m)` = depth) %>% 
  rename('Temp(°C)' = temp) %>% 
  rename('SO4(mM/Kg)' = SO4) %>%
  rename('Prok(cells/ml)' = prok) %>% 
  rename('TDN(μM)' = TDN) %>% 
  rename('TOC(μM)' = TOC) %>% 
  rename('Euk(cells/ml)' = eukCONC) %>%
  rename('Ca (mM/Kg)' = Ca) %>% 
  rename('Cl (mM/Kg)' = Cl) %>% 
  rename('Na (mM/Kg)' = Na) %>% 
  rename('K (mM/Kg)' = K) %>% 
  rename('Mg (mM/Kg)' = Mg) %>% 
  mutate(`18S` = as.character(`18S`)) %>% 
  mutate(depth_location = reorder(depth_location, -`Depth(m)`)) #order by depth_location using the depth column

variable_fill <- list(
  "Depth" = "Depth(m)",
  "Temp" = "Temp(°C)",
  "Euk Cell Count" = "Euk(cells/ml)",
  "Prok Cell Count" = "Prok(cells/ml)",
  "pH" = "pH",
  "TOC" = "TOC(μM)",
  "TDN" = "TDN(μM)",
  "SO4" = "SO4(mM/Kg)",
  "Ca" = "Ca (mM/Kg)",
  "Mg" = "Mg (mM/Kg)",
  "Cl" = "Cl (mM/Kg)",
  "Na" = "Na (mM/Kg)",
  "K" = "K (mM/Kg)",
  "18S" = "18S")

#Set scales
scale_fill <- list( 
  scale_fill_gradient(low = "#ADD", high = "#191970"), # depth
  scale_fill_gradient(low = "lightblue", high = "coral"), #temp
  scale_fill_gradient(low = "#d8d8f0", high = "#DA70D6"), # eukCONC
  scale_fill_gradient(low = "#d8d8f0", high = "#8B008B"), # prok
  scale_fill_gradient(low = "#B22", high = "#8A2BE2"), # pH
  scale_fill_gradient(low = "#efa", high = "#FF7F50"), # TOC
  scale_fill_gradient(low = "#e38aa6", high = "#D77"), # TDN
  scale_fill_gradient(low = "#a13356", high = "#8a1a3d"), # SO4
  scale_fill_gradient(low = "#b54a68", high = "#962748"), #Ca
  scale_fill_gradient(low = "#9ecae1", high = "#08519c"),  #Mg
  scale_fill_gradient(low = "#a1d99b", high = "#006d2c"),  #Cl
  scale_fill_gradient(low = "#bcbddc", high = "#54278f"),  #Na
  scale_fill_gradient(low = "#fdd0a2", high = "#d94801"),  #K
  scale_fill_manual(values = c("0" = "#F5FFFA", "1" = "#000000"))) #18S

names(scale_fill) <- names(variable_fill)

env_heatmap <- ggplot(env_data, aes(y = depth_location)) +
  purrr::imap(variable_fill, \(x, y) {
    list(
      geom_tile(
        aes(x = y, fill = .data[[x]]),
        color = "black",
        width = 0.95),
      scale_fill[[y]],
      ggnewscale::new_scale_fill())}) +
  theme_void() +
  theme(
    axis.text.y = element_text(hjust = 1),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_blank(),
    legend.position = "right",
  legend.box = "vertical",
  legend.direction = "horizontal") + 
  scale_x_discrete(limits = names(variable_fill))

ggsave("~/env_heatmap.png", plot = env_heatmap, width = 10, height = 10, dpi = 600, units = "in")


prok_location_order <- c("Background", "Water Column", "Above Vent", "Diffuse")
euk_location_order <- c("Background", "Water Column", "Diffuse")

prok_counts <- axial_metadata %>% 
  filter(prok >= 0) %>% 
  filter(!sample_type %in% c("Euphotic", "Oxycline")) %>% 
  mutate(sample_type = factor(sample_type,
                         levels = prok_location_order))

#Cell counts
prok_count <- ggplot(prok_counts, aes(x = sample_type, y = prok, color = location_name)) +
  geom_jitter(size = 3) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod",
    "Marker 113" = "violet",
    "Marker 33" = "purple")) +
  scale_y_log10() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  facet_grid(cols=vars(sample_type), space = "free", scales = "free") +
  labs(x = "", y = "Cells/ml (log)", color = NULL) +
  ggtitle("Prokaryotic Cell Abundance") + theme(legend.position = "none")

#Cell counts
euk_count <- axial_euk_cell_counts %>% 
  filter(!`Sample Type` %in% c("Euphotic", "Oxycline")) %>% 
  filter(!`Vent Field` %in% c("Coquille", "N3", "Bag  City")) %>% 
  mutate(`Sample Type` = factor(`Sample Type`,
                         levels = euk_location_order))
  
euk_plot <- ggplot(euk_count, aes(x = `Sample Type`,
         y = CELLS_PER_ML_total,
        color = `Vent Field`)) +
  geom_jitter(size = 3) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod",
    "Marker 113" = "violet",
    "Marker 33" = "purple"
  )) +
  scale_y_log10() +
  facet_grid(cols = vars(`Sample Type`), space = "free", scales = "free") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(x = "", y = NULL, color = NULL) +
  ggtitle("Eukaryotic Cell Abundance")

#library(patchwork)

cell_count_plot <- prok_count + euk_plot

ggsave("~/cell_count_plot.png", plot = cell_count_plot, width = 10, height = 5, dpi = 600, units = "in")

#TOC plot
unique(TOC_data$location_name)

TOC_data <- axial_metadata %>% 
  filter(!is.na(TOC)) 

TOC_plot <- ggplot(TOC_data, aes(x = TOC, y = sample_type, color = location_name, shape = sample_type)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(
    "Background" = 15,
    "Diffuse" = 17, 
    "Water Column" = 16,
    "Above Vent" = 16)) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod",
    "Marker 113" = "violet",
    "Marker 33" = "purple")) +
  theme_classic() + theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(shape = "none") +
  labs(x = "TOC (µM)", y = "") 

ggsave("~/TOC_plot.png", plot = TOC_plot, width = 10, height = 6, dpi = 600, units = "in")

```
#Alpha diversity
```{r}
#if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("phyloseq")

#library(phyloseq)

#Using axial_metadata_allsamples make a new DF and identify replicates

axial_allsamples_wReps <- axial_metadata_allsamples %>% 
  mutate(DepthCat_Location = paste(dive_ctd, filter_type, sample_name, location_name, depth_category, depth, sep = "-"))

#write.csv(axial_allsamples_wReps, file = "axial_allsamples_wReps.csv", row.names = FALSE)

#Re-upload file and combine replicate label to make unique sample identifier
axial_allsamples_wReps <- read_csv("R/Axial2023/axial_allsamples_wReps.csv") 

axial_allsamples_wReps <- axial_allsamples_wReps %>% 
  mutate(DepthCat_wReps = paste(DepthCat_Location, Rep, sep = "-"))

#Add DeptCat_wRepsn from metadata to ASV table
alpha_ASV <- ASV %>% 
  filter(FeatureID %in% samplestokeep_ASV$FeatureID) %>% 
  pivot_longer(cols = -FeatureID,
               names_to = "Sample", 
               values_to = "SEQUENCE_COUNT") %>% 
  filter(Sample %in% axial_allsamples_wReps$Sample) %>% 
  left_join(axial_allsamples_wReps %>% select(Sample, DepthCat_wReps), by = "Sample")

axial_allsamples_wReps <- axial_allsamples_wReps %>% column_to_rownames("DepthCat_wReps")

ASV_wide <- alpha_ASV %>%
  select(FeatureID, DepthCat_wReps, SEQUENCE_COUNT) %>%
  pivot_wider(
    names_from = DepthCat_wReps,
    values_from = SEQUENCE_COUNT,
    values_fill = 0)

#Phyloseq objects need to have row.names. Define row names from ASV column.
#Then transform into matrices. Metadata can be left as DF. 
asv_mat <- ASV_wide %>%  tibble::column_to_rownames("FeatureID") %>% as.matrix()
tax_mat <- tax %>% tibble::column_to_rownames("FeatureID") %>% as.matrix()

#Transform into phyloseq objects
ASV_phy = otu_table(asv_mat, taxa_are_rows = TRUE)
tax_phy = tax_table(tax_mat)
samples = sample_data(axial_allsamples_wReps)

#Final axial phyloseq object
Axial_phy <- phyloseq(ASV_phy, tax_phy, samples)

#Calculate alpha diversity
axial_alpha_div <- estimate_richness(
  Axial_phy,
  measures = c("Shannon", "Simpson", "Chao1"))

axial_alpha_div$SAMPLE <- sample_names(Axial_phy)

axial_alpha_div <- axial_alpha_div %>% 
  rownames_to_column("sample_dont_use") %>% 
  select(-sample_dont_use) %>% 
  separate(SAMPLE, into = c("Dive", "Filter", "DEPTH_CATEGORY", "Location_Name", "Depth", "Rep"),  
    sep = "-", 
    remove = TRUE)

sample_order <- c("Near Surface", "Background", "Caldera Basin Water", "Plume", "Above Vent", "Diffuse")

axial_alpha_div <- axial_alpha_div %>% 
  mutate(Depth = case_when(
    DEPTH_CATEGORY == "Caldera Basin Water" ~ "Caldera Basin Water",
    TRUE ~ Depth)) %>% 
  mutate(sample_type = case_when(
    Depth == "diffuse" ~ "Diffuse",
    Depth == "background" ~ "Background",
    Depth == "above vent" ~ "Above Vent",
    Depth == "euphotic" ~ "Near Surface",
    Depth == "oxycline" ~ "Near Surface",
    str_detect(Depth, "plume") ~ "Plume",
    str_detect(Depth, "30m") ~ "Plume",
    TRUE ~ Depth)) %>% 
  mutate(sample_type = factor(sample_type, levels = sample_order))

unique(axial_alpha_div$sample_type)

#axial_alpha_div_noreps <- axial_alpha_div %>% 
  mutate(SAMPLE = paste(Dive, Filter, DEPTH_CATEGORY, Location_Name, Depth, sep = "-")) %>%  
  group_by(SAMPLE) %>% 
  summarise(
    Shannon_mean = mean(Shannon, na.rm = TRUE),
    Simpson_mean = mean(Simpson, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop")

#Shannon  
shannon <- ggplot(
  axial_alpha_div,
  aes(x = sample_type, y = Shannon, color = Location_Name)) +
  geom_boxplot(aes(outlier_shape = NA)) +
  geom_jitter(width = 0.15, alpha = 0.7) +
  scale_color_manual(name = "Location", values = c(
    "ASHES" = "firebrick2",
    "Marker 33" = "purple",
    "Marker 113" = "blue",
    "Caldera Basin Water" = "green",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod")) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.title = element_blank(),
    legend.position = "none") +
  labs(title = "Shannon")

simpson <- ggplot(
  axial_alpha_div,
  aes(x = sample_type, y = Simpson, color = Location_Name)) +
  geom_boxplot(aes(outlier_shape = NA)) +
  geom_jitter(width = 0.15, alpha = 0.7) +
  scale_color_manual(name = "Location", values = c(
    "ASHES" = "firebrick2",
    "Marker 33" = "purple",
    "Marker 113" = "blue",
    "Caldera Basin Water" = "green",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod")) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.title = element_blank()) +
  labs(title = "Simpson")

chao <- ggplot(
  axial_alpha_div,
  aes(x = sample_type, y = se.chao1, color = Location_Name)) +
  geom_boxplot(aes(outlier_shape = NA)) +
  geom_jitter(width = 0.15, alpha = 0.7) +
  scale_color_manual(name = "Location", values = c(
    "ASHES" = "firebrick2",
    "Marker 33" = "purple",
    "Marker 113" = "blue",
    "Caldera Basin Water" = "green",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod")) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank(),
    legend.position = "none") +
  labs(title = "Chao1")

#library(patchwork)

alpha_patch <- shannon/simpson/chao
alpha_patch

ggsave("~/alpha_patch.png", plot = alpha_patch, width = 8, height = 10, units = "in")
```

#NMDS and PERMANOVA
```{r}
#NMDS without collapsed replicates

#Compute Bray-Curtis and run NMDS

NMDS_input <- tax_asv_metadata %>% 
  filter(!depth_category %in% c("Euphotic", "Oxycline")) %>%
  select(FeatureID, Sample, SEQUENCE_COUNT) %>% 
  pivot_wider(names_from = FeatureID,
              values_from = SEQUENCE_COUNT,
              values_fill = 0) %>% 
  column_to_rownames("Sample") %>% 
  as.matrix() 

#Compute Bray-Curtis and run NMDS
set.seed(123)
axial_nmds <- metaMDS(NMDS_input, distance = "bray") ##Stress=0.14

# Extract NMDS coordinates for samples
nmds_scores <- as.data.frame(scores(axial_nmds, display = "sites"))

# Add Sample names as a column
nmds_scores <- scores(axial_nmds, display = "sites") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample")

NMDS_metadata <- tax_asv_metadata %>% 
  left_join(nmds_scores, by = "Sample") %>% 
  mutate(sample_type = case_when(
    sample_name == "Caldera Basin Water" ~ "Caldera Basin Water",
    TRUE ~ sample_type)) %>% 
  filter(!sample_type %in% c("Euphotic", "Oxycline"))

unique(NMDS_metadata$sample_type)
#Plot NMDS
#install.packages("ggnewscale")
#library(ggnewscale)

NMDS_plot <- ggplot(NMDS_metadata, aes(x = NMDS1, y = NMDS2)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +

  # WATER COLUMN FIRST
  geom_point(
    data = filter(NMDS_metadata, sample_type == "Water Column"),
    aes(shape = sample_type,
        fill = depth),
    size = 3) +

  scale_fill_gradient(
    high = "darkblue",
    low = "lightblue",
    name = "Water Column Depth (m)",
    guide = guide_colorbar(reverse = TRUE)) +

  ggnewscale::new_scale_fill() +

  # NON-WATER COLUMN SECOND
geom_point(
    data = filter(NMDS_metadata, sample_type != "Water Column"),
    aes(shape = sample_type,
        fill = sample_name),
    size = 3) +

  scale_fill_manual(
    name = "Sample Location",
    values = c(
      "Anemone" = "lightgoldenrod",
      "Gollum" = "orange",
      "ASHES" = "firebrick2",
      "International District" = "darkcyan",
      "Diva" = "darkcyan",
      "Background" = "darkblue",
      "Caldera Basin Water" = "darkblue",
      "Marker 33" = "mediumpurple",
      "Marker 113" = "mistyrose3")) +

  scale_shape_manual(name = "Sample Type",
    values = c(
    "Above Vent" = 25,
    "Diffuse" = 24,
    "Water Column" = 21,
    "Background" = 22,
    "Caldera Basin Water" = 23)) +
  guides(fill = guide_legend(override.aes = list(shape = 21, stroke = 0.5)))


ggsave("~/NMDS_plot.png", plot = NMDS_plot, width = 6, height = 6, dpi = 600, units = "in")

#Run PERMANOVA. 
#1. ASV table where rows = samples and columns = ASVs. This is NMDS_input
#2. Metadata table where row = sample. This looks like NMDS_metadata. Calculate BC. 

adonis_ASV <- NMDS_input %>% 
  as.data.frame()

# Metadata with Sample as rownames
adonis_metadata <- axial_metadata %>%
  filter(Sample %in% samplestokeep_ASV$Sample) %>% 
  column_to_rownames("Sample")

# Ensure identical order 
adonis_ASV <- adonis_ASV[rownames(adonis_metadata), ]

# Check alignment
all(rownames(adonis_ASV) == rownames(adonis_metadata)) #TRUE

setdiff(rownames(adonis_ASV), rownames(adonis_metadata))
setdiff(rownames(adonis_metadata), rownames(adonis_ASV))

# Bray-Curtis distance
bray <- vegdist(adonis_ASV, method = "bray")

# PERMANOVA
adonis_results <- adonis2(
  bray ~ sample_type,
  data = adonis_metadata,
  permutations = 9999)

```


#Jaccard distance decay and dendogram
```{r}
library("phyloseq")

#Format ASV and tax table
asv_matrix <- asv_wide %>% 
  as.data.frame() %>% 
  rownames_to_column("depth_location") %>% 
  pivot_longer(cols = -depth_location,
               names_to = "FeatureID",
               values_to = "mean_seq") %>% 
  separate(depth_location,
    into = c("location","site","depth"),  
    sep = "-") %>% 
  mutate(SampleID = paste(location, site, depth, sep = "-")) %>% 
  mutate(SampleID = case_when(
  str_detect(SampleID, "Background-1200") ~ "Background-1200",
   str_detect(SampleID, "Background-1300") ~ "Background-1300",
   str_detect(SampleID, "Background-1400") ~ "Background-1400",
  str_detect(SampleID, "Caldera") ~ "Caldera Basin Water-1535",
    TRUE ~ SampleID)) %>% 
  select(-location, -site, -depth) %>% 
  group_by(FeatureID, SampleID) %>%  
  summarise(seq_sum = sum(mean_seq)) %>% 
  pivot_wider(names_from = SampleID, 
              values_from = seq_sum, values_fill = 0) %>% 
  column_to_rownames("FeatureID") %>% 
  as.matrix()

tax_matrix <- tax %>% 
  separate(Taxon, into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>% select(-Consensus) %>% column_to_rownames("FeatureID") %>% 
  as.matrix()

tax_table <- tax_table(tax_matrix)
ASV_table <- otu_table(asv_matrix, taxa_are_rows = TRUE)

physeq_object <- merge_phyloseq(ASV_table, tax_table, axial_metadata)

dist <- phyloseq::distance(physeq_object, method = "jaccard")

rownames(as.data.frame(as.matrix(dist)))

dist <- as.data.frame(as.matrix(dist))

dist_scores_df <- dist %>% 
  rownames_to_column(var = "SampleA") %>% 
  pivot_longer(cols = -SampleA, 
               names_to = "SampleB")

metadata <- axial_metadata %>% 
  filter(sample_type != "High Temp Vent") %>% 
  select(depth_location, lat, long, depth) %>% distinct() 
  

dist_df2 <- dist_scores_df %>%
  
  left_join(metadata %>% 
              select(depth_location, lat, long, depth),
            by = c("SampleA" = "depth_location")) %>%
  rename(latA = lat,
         longA = long,
         depthA = depth) %>%
  
  left_join(metadata %>% 
              select(depth_location, lat, long, depth),
            by = c("SampleB" = "depth_location")) %>%
  rename(latB = lat,
         longB = long,
         depthB = depth)

#install.packages("geosphere")
#library(geosphere)

dist_df2 <- dist_df2 %>%
  mutate(
    horizontal_distance = distHaversine(
      cbind(longA, latA),
      cbind(longB, latB)),
    z_dist = abs(depthA - depthB),
    distance_3d = sqrt(horizontal_distance^2 + z_dist^2)) %>% 
  mutate(samples_combined = paste(SampleA, SampleB, sep = "-")) %>%
  mutate(bin = cut(distance_3d, breaks = 10)) 

unique(dist_df2$distance_3d)


distance_decay_GRC <- ggplot(dist_df2, aes(x = distance_3d, y = value)) +
  geom_jitter() +
  theme_linedraw() +
  scale_y_continuous(limits = c(0.4,1)) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black", face = "bold", size = 12),
        axis.text = element_text(color = "black", face = "bold", size = 12),
        axis.title = element_text(color = "black", face = "bold", size = 12)) +
  labs(y = "Jaccard Dissimilarity", x = "Euclidean Distance (m)") #+
  geom_text(aes(label = samples_combined))
  
ggsave("~/distance_decay_GRC.png", plot = distance_decay_GRC, width = 6, height = 6, dpi = 600, units = "in")

#install.packages("ggdendro")
library(ggdendro)

dist_plot <- phyloseq::distance(
  physeq_object,
  method = "jaccard",
  binary = TRUE)

hc <- hclust(dist_plot, method = "average")

#plot(hc)

cluster_jacc <- hclust(dist_plot, method = "average")

dendro_plot_df <- ggdendro::dendro_data(as.dendrogram(cluster_jacc), type = "rectangle")
label_dendro_order <- as.character(dendro_plot_df$labels$label)

ggdendro_GRC <- ggplot(segment(dendro_plot_df)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  coord_flip() + 
  scale_y_reverse(expand = c(0.2, 0.5), breaks = c(0, 0.2, 0.4, 0.6, 0.8)) + 
  geom_text(
  aes(x = x, y = y, label = label, angle = 0, hjust = 0),
  data = label(dendro_plot_df),
  size = 5) + 
  theme_dendro() + 
  labs(y = "Jaccard Dissimilarity") + 
theme(
  axis.title.x = element_text(size = 20, color = "black"),
  axis.text.x = element_text(color = "black", size = 20),
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  axis.line.x = element_line(color = "#252525"),
  axis.ticks.x = element_line())



ggsave("~/ggdendro_GRC.png", plot = ggdendro_GRC, width = 12, height = 8, dpi = 600, units = "in")
head(label_dendro_order)

```

#Stacked bar plot in dendogram order
```{r}
#Keep ASVs that only have 50 or more sequences and are found in two or more samples
asv_keep <- ASV %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "sample",
    values_to = "SEQUENCE_COUNT") %>%
  group_by(FeatureID) %>%
  summarise(
    total_reads = sum(SEQUENCE_COUNT),
    prevalence = sum(SEQUENCE_COUNT > 0),
    .groups = "drop") %>%
  filter(total_reads >= 50, prevalence >= 2)

samplestokeep_ASV <- ASV %>% 
  filter(FeatureID %in% asv_keep$FeatureID) %>% 
  pivot_longer(
    cols = -c(FeatureID),
    names_to = "Sample",
    values_to = "SEQUENCE_COUNT") %>%
  filter(Sample %in% axial_metadata$Sample) 

table_asv <-  samplestokeep_ASV %>% 
  left_join(axial_metadata, by = "Sample") %>% 
  left_join(tax, by = "FeatureID") %>% 
  separate(
    col = Taxon,
    into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Genus", "Species"),
    sep = ";")

alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")

#Curate taxonomy
#library(dplyr)
#library(stringr)
#library(purrr)

collapsed_reps_ASV <- tax_asv_metadata %>% 
  filter(depth_category != "euphotic") %>% 
  filter(depth_category!= "oxycline") %>% 
  mutate(depth_location = case_when(
  str_detect(depth_location, "Background-1200") ~ "Background-1200",
   str_detect(depth_location, "Background-1300") ~ "Background-1300",
   str_detect(depth_location, "Background-1400") ~ "Background-1400",
  str_detect(depth_location, "Caldera") ~ "Caldera Basin Water-1535",
    TRUE ~ depth_location)) %>% 
  group_by(taxa, depth_location) %>% 
  summarise(
    seq_sum = sum(SEQUENCE_COUNT, na.rm = TRUE),
    .groups = "drop") %>% 
  group_by(depth_location) %>% 
   mutate(sample_seq_total = sum(seq_sum),
         taxa_relAbun = seq_sum / sample_seq_total) %>% 
  ungroup() 

collapsed_reps_ASV <- collapsed_reps_ASV %>% mutate(
    depth_location = factor(depth_location,
                         levels = label_dendro_order))
  

taxa <- c("Alveolata-Dinoflagellata" = "cornflowerblue","Rhizaria-Radiolaria" = 
            "coral4","Alveolata-Ciliophora" = "coral", 
          "Stramenopiles-Gyrista" = "chocolate","Rhizaria-Cercozoa" = "cadetblue","Stramenopiles-Bigyra" = "aquamarine","Archaeplastida" = "firebrick4", 
          "Discosea-Unknown" = "plum4","Stramenopiles-Other" = "lightsalmon", "Amorphea" = "darkolivegreen3","Haptista" = "gold","Amoebozoa" = "lightgoldenrod",
          "Unknown Eukaryota" = "mistyrose","Telonemia" = "rosybrown","Alveolata-Other" = "peachpuff4","Cryptista" = "coral3","CRuMS" = "darkcyan",
          "Excavata" = "orchid","Provora-Nibbleridia" = "powderblue")


GRC_stackedbar <- ggplot(collapsed_reps_ASV, aes(y = depth_location, x = taxa_relAbun, fill = taxa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = taxa) +
  theme_minimal(base_size = 12) +
  theme(
    # Text
    axis.text.x = element_text(face = "bold", color = "black", size = 14),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title = element_text(color = "black", face = "bold"),
    strip.text = element_text(color = "black", face = "bold", size = 12),
    legend.text = element_text(color = "black", size = 14),
    
    # Backgrounds
     plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    strip.background = element_rect(fill = "gray", color = NA),
    
    # Gridlines
    panel.grid = element_blank(),
    
    # Legend
    legend.position = "right",
    legend.title = element_blank()) +
  labs(
    title = "",
    x = "",
    y = "") +
  theme(
  legend.text = element_text(size = 16)) +
  guides(fill = guide_legend(title = NULL))

ggsave("~/GRC_stackedbar.png", plot = GRC_stackedbar, width = 6, height = 8, dpi = 600, units = "in")

library(patchwork)

combined_plot <- ggdendro_GRC + GRC_stackedbar +
  plot_layout(widths = c(12, 3))

ggsave("~/combined_plot.png", plot = combined_plot, width = 12, height = 8, dpi = 600, units = "in")
```

#Chord plot
Chord diagram shows the flow between a set of entities (nodes)
```{r}
#install.packages("circlize")
#library(circlize)
#library(RColorBrewer)

#Format ASV table
chord_df <- tax_asv_metadata %>%
  group_by(sample_type, taxonomy_collapse) %>%
  summarise(abundance_avg = mean(SEQUENCE_COUNT), .groups = "drop") %>%
  filter(
    !is.na(sample_type),
    !is.na(taxonomy_collapse),
    abundance_avg > 0) %>%
  mutate(
    sample_type = as.character(sample_type),
    taxonomy_collapse = as.character(taxonomy_collapse))

top_taxa <- chord_df %>%
  group_by(taxonomy_collapse) %>%
  summarise(total = sum(abundance_avg), .groups = "drop") %>%
  slice_max(total, n = 45) %>%
  pull(taxonomy_collapse)

chord_df_top45 <- chord_df %>%
  filter(taxonomy_collapse %in% top_taxa)

taxa_lookup <- data.frame(
  taxonomy_collapse = sort(unique(chord_df_top45$taxonomy_collapse))
) %>%
  mutate(tax_id = as.character(row_number()))

chord_df_plot <- chord_df_top45 %>%
  left_join(taxa_lookup, by = "taxonomy_collapse")

tax_colors <- setNames(
  colorRampPalette(brewer.pal(12, "Set3"))(nrow(taxa_lookup)),
  taxa_lookup$tax_id)

#Build alist of every unique category that will appear in the chord diagram
all_sectors <- unique(c(
  chord_df_top45$sample_type,
  chord_df_top45$taxonomy_collapse))

sample_colors <- setNames(
  rep("grey80", length(unique(chord_df_plot$sample_type))),
  unique(chord_df_plot$sample_type))

grid.col <- c(sample_colors, tax_colors)

plot_df <- chord_df_plot %>%
  transmute(
    sample_type,
    taxonomy = tax_id,
    value = abundance_avg)
#Run below as one chunk or it wont work

svg("chord_diagram.svg")

circos.clear()

chordDiagram(
  plot_df,
  grid.col = grid.col,
  col = tax_colors[plot_df$taxonomy],
  transparency = 0.4)

dev.off() 
##

legend_df <- taxa_lookup %>%
  arrange(as.numeric(tax_id)) %>%
  mutate(tax_id = factor(tax_id, levels = tax_id)) %>%
  mutate(color = tax_colors[tax_id]) %>% 
  mutate(tax_wNumber = paste(tax_id, taxonomy_collapse, sep = "-")) %>% 
  arrange(as.numeric(tax_id)) %>%
  mutate(tax_wNumber = factor(tax_wNumber, levels = tax_wNumber))
  

chord_legend <- ggplot(legend_df, aes(x = tax_wNumber, y = 1, fill = tax_id)) +
  geom_tile() +
  scale_fill_manual(name = "Taxonomy", values = tax_colors, breaks = legend_df$tax_id) +
  theme_void() +
  theme(legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("chord_legend.png", chord_legend, height = 8, width = 10, dpi = 600)

```

#Chord plot at the species level for strams (Bigyra, gyrista), rhizaria (radiolaria), haptista
```{r}
#install.packages("circlize")
library(circlize)
library(RColorBrewer)

#Filter for taxa of interest

SRH <- tax_asv_metadata %>% 
  filter(taxa %in% c("Rhizaria-Radiolaria", "Stramenopiles-Bigyra", "Stramenopiles-Gyrista", "Haptista")) %>% 
  mutate(Species = case_when(
  (is.na(Genus) | Genus == "Unknown") & is.na(Species) ~ "Unknown",
  !is.na(Genus) & Genus != "Unknown" & is.na(Species) ~ paste0(Genus, "-Unknown"),
   str_detect(Species, "_X") ~ str_replace(Species, "_X+$", "-Unknown"),
  TRUE ~ Species)) %>% 
  mutate(taxa_species = paste(taxa, Species, sep = "-"))

unique(SRH$taxa_species)

#Format ASV table
chord_df_SRH <- SRH %>%
  group_by(sample_type, taxa_species) %>%
  summarise(abundance_avg = mean(SEQUENCE_COUNT), .groups = "drop") %>%
  filter(abundance_avg > 0) %>% 
  mutate(
    sample_type = as.character(sample_type),
    taxa_species = as.character(taxa_species)) %>% 
  ungroup()

#Pull out top taxa for each sample category
top20_abovevent <- chord_df_SRH %>% 
  filter(sample_type == "Above Vent") %>% 
  group_by(sample_type, taxa_species) %>%
  summarise(
    overall_abundance = mean(abundance_avg),
    .groups = "drop") %>%
  slice_max(
    overall_abundance,
    n = 20)

top20_diffuse <- chord_df_SRH %>% 
  filter(sample_type == "Diffuse") %>% 
  group_by(sample_type, taxa_species) %>%
  summarise(
    overall_abundance = mean(abundance_avg),
    .groups = "drop") %>%
  slice_max(
    overall_abundance,
    n = 20)

top20_background <- chord_df_SRH %>% 
  filter(sample_type == "Background") %>% 
  group_by(sample_type, taxa_species) %>%
  summarise(
    overall_abundance = mean(abundance_avg),
    .groups = "drop") %>%
  slice_max(
    overall_abundance,
    n = 20)

top20_watercolumn <- chord_df_SRH %>% 
  filter(sample_type == "Water Column") %>% 
  group_by(sample_type, taxa_species) %>%
  summarise(
    overall_abundance = mean(abundance_avg),
    .groups = "drop") %>%
  slice_max(
    overall_abundance,
    n = 20)

top20_CBW <- chord_df_SRH %>% 
  filter(sample_type == "Caldera Basin Water") %>% 
  group_by(sample_type, taxa_species) %>%
  summarise(
    overall_abundance = mean(abundance_avg),
    .groups = "drop") %>%
  slice_max(
    overall_abundance,
    n = 20)

top20_all <- bind_rows(
  top20_abovevent,
  top20_diffuse,
  top20_background,
  top20_watercolumn,
  top20_CBW)

n_distinct(top20_all$taxa_species) #Went from 121 total to 42

taxa_lookup_top20_all <- data.frame(
    taxa_species = sort(unique(top20_all$taxa_species))) %>%
    mutate(tax_id = as.character(row_number()))

chord_df_plot <- top20_all %>%
    left_join(taxa_lookup_top20_all, by = "taxa_species")

tax_colors <- setNames(
    colorRampPalette(brewer.pal(12,"Set3"))(nrow(taxa_lookup_top20_all)),
    taxa_lookup_top20_all$tax_id)

#Build alist of every unique category that will appear in the chord diagram
sample_colors <- setNames(
  rep("grey80", length(unique(top20_all$sample_type))),
  unique(top20_all$sample_type))

grid.col <- c(sample_colors, tax_colors)

plot_df <- chord_df_plot %>%
    transmute(
        sample_type,
        taxonomy = tax_id,
        value = overall_abundance)

n_distinct(top30_all$taxa_species)

#Run below as one chunk or it wont work
svg("chord_diagram_SRH.svg")

circos.clear()

chordDiagram(
  plot_df,
  grid.col = grid.col,
  col = tax_colors[plot_df$taxonomy],
  transparency = 0.4)

dev.off() 
##

legend_df <- taxa_lookup_top20_all %>%
  arrange(as.numeric(tax_id)) %>%
  mutate(tax_id = factor(tax_id, levels = tax_id)) %>%
  mutate(color = tax_colors[tax_id]) %>% 
  mutate(tax_wNumber = paste(tax_id, taxa_species, sep = "-")) %>% 
  arrange(as.numeric(tax_id)) %>%
  mutate(tax_wNumber = factor(tax_wNumber, levels = tax_wNumber))
  

chord_legend_SRH <- ggplot(legend_df, aes(x = tax_wNumber, y = 1, fill = tax_id)) +
  geom_tile() +
  scale_fill_manual(name = "Taxonomy", values = tax_colors, breaks = legend_df$tax_id) +
  theme_void() +
  theme(legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("chord_legend_SRH.png", chord_legend_SRH, height = 8, width = 10, dpi = 600)
```

#chord plot for strams, haptista and radiolaria at the order level 
```{r}
SRH_order <- tax_asv_metadata %>% 
  filter(taxa %in% c("Rhizaria-Radiolaria", "Stramenopiles-Bigyra", "Stramenopiles-Gyrista", "Haptista")) %>% 
  mutate(taxa_order = paste(taxa, Order, sep = "-"))

chord_order_SRH <- SRH_order %>%
  group_by(sample_type, taxa_order) %>%
  summarise(abundance_avg = mean(SEQUENCE_COUNT), .groups = "drop") %>%
  filter(abundance_avg > 0) %>% 
  mutate(
    sample_type = as.character(sample_type),
    taxa_order = as.character(taxa_order)) %>% 
  ungroup()

n_distinct(chord_df_SRH$taxa_order) #30 unique IDs

diffuse <- chord_order_SRH %>% 
  filter(sample_type == "Diffuse") 

sum(diffuse$abundance_avg, na.rm = TRUE) #72211 sequences for diffuse

sum()

taxa_lookup_order <- data.frame(
    taxa_order = sort(unique(chord_order_SRH$taxa_order))) %>%
    mutate(tax_id = as.character(row_number()))

chord_order_plot <- chord_order_SRH %>%
    left_join(taxa_lookup_order, by = "taxa_order")

tax_colors <- setNames(
    colorRampPalette(brewer.pal(12,"Set3"))(nrow(taxa_lookup_order)),
    taxa_lookup_order$tax_id)

#Build alist of every unique category that will appear in the chord diagram
sample_colors <- setNames(
  rep("grey80", length(unique(chord_order_plot$sample_type))),
  unique(chord_order_plot$sample_type))

grid.col <- c(sample_colors, tax_colors)

plot_order_df <- chord_order_plot %>%
    transmute(
        sample_type,
        taxonomy = tax_id,
        value = abundance_avg)

#Run below as one chunk or it wont work
svg("chord_SRH_order.svg")

circos.clear()

chordDiagram(
  plot_df,
  grid.col = grid.col,
  col = tax_colors[plot_order_df$taxonomy],
  transparency = 0.4)

dev.off() 
##

legend_order <- taxa_lookup_order %>%
  arrange(as.numeric(tax_id)) %>%
  mutate(tax_id = factor(tax_id, levels = tax_id)) %>%
  mutate(color = tax_colors[tax_id]) %>% 
  mutate(tax_wNumber = paste(tax_id, taxa_order, sep = "-")) %>% 
  arrange(as.numeric(tax_id)) %>%
  mutate(tax_wNumber = factor(tax_wNumber, levels = tax_wNumber))
  

chord_legend_order <- ggplot(legend_order, aes(x = tax_wNumber, y = 1, fill = tax_id)) +
  geom_tile() +
  scale_fill_manual(name = "Taxonomy", values = tax_colors, breaks = legend_df$tax_id) +
  theme_void() +
  theme(legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("chord_legend_order.png", chord_legend_order, height = 8, width = 10, dpi = 600)

```


#RDA
```{r}
#Generate a matrix of numerical data
numerical <- axial_metadata %>%
  select(depth_location, depth, temp, sample_name, prok) %>% 
  group_by(depth_location) %>%
  summarise(
    depth = mean(depth),
    prok = mean(prok),
    temp = mean(temp),
    .groups = "drop"
  )  %>% 
  filter(!is.na(prok))

#Make a categorical data table 
categorical <- axial_metadata %>% 
  select(sample_type, location_name, depth_category, 
         sample_name, depth_location) %>% filter(depth_location %in% numerical$depth_location) %>% 
  distinct()
  
numerical <- numerical %>% column_to_rownames("depth_location")

standardize_numerical <- decostand(numerical, MARGIN = 2, method = "standardize")

# Combine normalized metadata with categorical data
standardize_numerical <- standardize_numerical %>% 
  rownames_to_column("depth_location")

metadata_normalized <- categorical %>% 
  right_join(as.data.frame(standardize_numerical)) %>% 
  column_to_rownames(var = "depth_location")
  
seq_wide <- tax_asv_metadata %>% 
  select(FeatureID, SEQUENCE_COUNT, depth_location) %>% 
  group_by(depth_location, FeatureID) %>% 
  summarise(seq_avg = mean(SEQUENCE_COUNT), .groups = "drop") %>% 
  pivot_wider(names_from = FeatureID,
              values_from = seq_avg,
              values_fill = 0) %>% 
  column_to_rownames("depth_location") %>%  as.matrix()


clr_asv_matrix <- data.frame(compositions::clr(seq_wide))

asv_list <- colnames(seq_wide)

# asv_list
meta_list <- rownames(metadata_normalized)
clr_list <- rownames(clr_asv_matrix)
setdiff(meta_list, clr_list)

shared_samples <- intersect(
  rownames(clr_asv_matrix),
  rownames(metadata_normalized))

clr_asv_matrix <- clr_asv_matrix[shared_samples, ]
metadata_normalized <- metadata_normalized[shared_samples, ]

all(rownames(clr_asv_matrix) == rownames(metadata_normalized))
#Large f-ratio means there is a lot of between group variance compared to within-group variance (independent variables have a greater influence on depdendent variables). Smaller f-ratios indicate the groups are similar to one another. 

rda_clr <- rda(clr_asv_matrix ~ ., data = metadata_normalized)

summary(rda_clr)

anova(rda_clr, by = "term", permutations = 9999)

```

#RDA without vent samples - didnt work?
```{r}
#Generate a matrix of numerical data
numerical_novent <- axial_metadata %>%
  select(sample_type, depth_location, depth, temp, sample_name, prok) %>% 
  filter(!sample_type %in% c("Euphotic", "Oxycline", "High Temp Vent", "Diffuse")) %>% 
  select(-sample_type) %>% 
  group_by(depth_location) %>%
  summarise(
    depth = mean(depth),
    prok = mean(prok),
    temp = mean(temp),
    .groups = "drop"
  )  %>% 
  filter(!is.na(prok))


#Make a categorical data table 
categorical_novent <- axial_metadata %>% 
  select(location_name, depth_category, 
         sample_name, depth_location) %>% filter(depth_location %in% numerical_novent$depth_location) %>% 
  distinct()
  
numerical_novent <- numerical_novent %>% column_to_rownames("depth_location")

standardize_numerical_novent <- decostand(numerical_novent, MARGIN = 2, method = "standardize")

# Combine normalized metadata with categorical data
standardize_numerical_novent <- standardize_numerical_novent %>% 
  rownames_to_column("depth_location")

metadata_normalized_novent <- categorical_novent %>% 
  right_join(as.data.frame(standardize_numerical_novent)) %>% 
  column_to_rownames(var = "depth_location")
  
seq_wide_novent <- tax_asv_metadata %>% 
  select(FeatureID, SEQUENCE_COUNT, depth_location) %>% 
  group_by(depth_location, FeatureID) %>% 
  summarise(seq_avg = mean(SEQUENCE_COUNT), .groups = "drop") %>% 
  pivot_wider(names_from = FeatureID,
              values_from = seq_avg,
              values_fill = 0) %>% 
  column_to_rownames("depth_location") %>%  as.matrix()


clr_asv_matrix_novent <- data.frame(compositions::clr(seq_wide_novent))

asv_list_novent <- colnames(seq_wide_novent)

# asv_list
meta_list_novent <- rownames(metadata_normalized_novent)
clr_list_novent <- rownames(clr_asv_matrix_novent)
setdiff(meta_list_novent, clr_list_novent)

shared_samples_novent <- intersect(
  rownames(clr_asv_matrix_novent),
  rownames(metadata_normalized_novent))

clr_asv_matrix_novent <- clr_asv_matrix_novent[shared_samples_novent, ]
metadata_normalized_novent <- metadata_normalized_novent[shared_samples_novent, ]

all(rownames(clr_asv_matrix_novent) == rownames(metadata_normalized_novent))
#Large f-ratio means there is a lot of between group variance compared to within-group variance (independent variables have a greater influence on depdendent variables). Smaller f-ratios indicate the groups are similar to one another. 

rda_clr_novent <- rda(clr_asv_matrix_novent ~ ., data = metadata_normalized_novent)

summary(rda_clr_novent)

anova_novent <- anova.cca(rda_clr_novent, by = "term")

vif.cca(rda_clr_novent)

```

#CLR dendrogram and heatmap
```{r}
#library(compositions)

CLR_df <- axial_metadata_nosurface

clr_asv <- tax_asv_metadata %>% 
  select(FeatureID, Sample, SEQUENCE_COUNT)

clr_matrix <- clr_asv %>%
  pivot_wider(
    id_cols = FeatureID,
    names_from = Sample,
    values_from = SEQUENCE_COUNT,
    values_fill = 0
  ) %>%
  column_to_rownames("FeatureID") %>%
  as.matrix()

clr_matrix <- clr(t(clr_matrix))
clr_matrix <- t(clr_matrix)

sample_meta <- CLR_df %>%
  distinct(Sample, sample_type)

tax_meta <- tax_asv_metadata %>%
  filter(FeatureID %in% clr_asv$FeatureID) %>% 
  distinct(FeatureID, taxa, Species) %>% 
  mutate(Species = case_when(
      is.na(Species) ~ "Unknown",
      TRUE ~ Species)) 

sample_type_order <- c("Background", "Water Column", "Caldera Basin Water",
                       "Above Vent", "Diffuse")

clr_tax_asv <- clr_matrix %>%
  as.data.frame() %>%
  rownames_to_column("FeatureID") %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "Sample",
    values_to = "clr"
  ) %>%
  left_join(sample_meta, by = "Sample") %>%
  left_join(tax_meta, by = "FeatureID") %>%
  group_by(taxa, Species, sample_type) %>%
  summarise(avg_clr = mean(clr), .groups = "drop") %>% 
  mutate(sample_type = factor(sample_type, levels = sample_type_order)) %>% 
  arrange(taxa, Species) %>% 
  mutate(Species = factor(Species, levels = unique(Species)))

clr_tax_asv <- clr_tax_asv %>% 
  filter(taxa %in% c("Stramenopiles-Bigyra", "Stramenopiles-Gyrista", 
                     "Haptista", "Rhizaria-Radiolaria"))

heatmap_revised <- ggplot(clr_tax_asv, aes(x = sample_type, y = Species, fill = avg_clr)) +
  geom_tile() +
  facet_grid(taxa ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/heatmap_revised.png", plot = heatmap_revised, width = 8, height = 12, dpi = 600, units = "in")

#library(pheatmap)

heatmap_df <- clr_tax_asv %>%
  unite("taxa_species", taxa, Species, sep = " - ") %>%
  pivot_wider(
    names_from = sample_type,
    values_from = avg_clr
  ) %>%
  column_to_rownames("taxa_species")

dendro_revised <- pheatmap(
  as.matrix(heatmap_df),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  scale = "none",
  color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(100))

ggsave("~/R/Axial2023/Figures/dendro_revised.png", plot = dendro_revised, width = 12, height = 12, dpi = 600, units = "in")

mast <- tax_asv_metadata %>% 
  filter(str_detect(Species, "MAST"))

```

#Whisker plot
```{r}
major_taxa <- tax_asv_metadata %>%
  filter(taxa %in% c(
    "Stramenopiles-Bigyra",
    "Stramenopiles-Gyrista",
    "Haptista",
    "Rhizaria-Radiolaria"
  )) %>%
  group_by(Sample, taxa) %>%
  summarise(
    sequences = sum(SEQUENCE_COUNT),
    .groups = "drop"
  ) %>%
  left_join(
    axial_metadata_nosurface %>%
      select(Sample, sample_type),
    by = "Sample"
  ) %>%
  group_by(Sample) %>%
  mutate(
    relative_abundance = sequences / sum(sequences)
  ) %>%
  ungroup()

whisker <- ggplot(
  major_taxa,
  aes(x = sample_type, y = relative_abundance)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7) +
  facet_wrap(~ taxa) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/whisker_plot.png", plot = whisker, width = 8, height = 10, dpi = 600, units = "in")

```

#Remake and keep sample names
```{r}
#library(compositions)

CLR_df <- axial_metadata_nosurface

clr_asv <- tax_asv_metadata %>% 
  select(FeatureID, Sample, SEQUENCE_COUNT)

clr_matrix <- clr_asv %>%
  pivot_wider(
    id_cols = FeatureID,
    names_from = Sample,
    values_from = SEQUENCE_COUNT,
    values_fill = 0
  ) %>%
  column_to_rownames("FeatureID") %>%
  as.matrix()

clr_matrix <- clr(t(clr_matrix))
clr_matrix <- t(clr_matrix)

sample_meta <- CLR_df %>%
  distinct(Sample, sample_type, depth_location, sample_name)

tax_meta <- tax_asv_metadata %>%
  filter(FeatureID %in% clr_asv$FeatureID) %>% 
  distinct(FeatureID, taxa, Species) %>% 
  mutate(Species = case_when(
      is.na(Species) ~ "Unknown",
      TRUE ~ Species)) 

clr_tax_asv <- clr_matrix %>%
  as.data.frame() %>%
  rownames_to_column("FeatureID") %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "Sample",
    values_to = "clr"
  ) %>%
  left_join(sample_meta, by = "Sample") %>%
  left_join(tax_meta, by = "FeatureID") %>%
  group_by(taxa, Species, depth_location, sample_name) %>% #Average replicates
  summarise(avg_clr = mean(clr), .groups = "drop") %>%
  arrange(taxa, Species) %>% 
  mutate(Species = factor(Species, levels = unique(Species)))

clr_tax_asv <- clr_tax_asv %>% 
  filter(taxa %in% c("Stramenopiles-Bigyra", "Stramenopiles-Gyrista", 
                     "Haptista", "Rhizaria-Radiolaria"))

clr_heatmap_2 <- ggplot(clr_tax_asv, aes(x = depth_location, y = Species, fill = avg_clr)) +
  geom_tile() +
 facet_grid(
  rows = vars(taxa),
  cols = vars(sample_name),
  scales = "free",
  space = "free") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/clr_heatmap_2.png", plot = clr_heatmap_2, width = 12, height = 12, dpi = 600, units = "in")

#library(pheatmap)

heatmap_df <- clr_tax_asv %>%
  unite("taxa_species", taxa, Species, sep = " - ") %>%
  pivot_wider(
    names_from = sample_type,
    values_from = avg_clr
  ) %>%
  column_to_rownames("taxa_species")

dendro_revised <- pheatmap(
  as.matrix(heatmap_df),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  scale = "none",
  color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(100))

ggsave("~/R/Axial2023/Figures/dendro_revised.png", plot = dendro_revised, width = 12, height = 12, dpi = 600, units = "in")

mast <- tax_asv_metadata %>% 
  filter(str_detect(Species, "MAST"))

```

#Whisker plot
```{r}
major_taxa <- tax_asv_metadata %>%
  filter(taxa %in% c(
    "Stramenopiles-Bigyra",
    "Stramenopiles-Gyrista",
    "Haptista",
    "Rhizaria-Radiolaria"
  )) %>%
  group_by(Sample, taxa) %>%
  summarise(
    sequences = sum(SEQUENCE_COUNT),
    .groups = "drop"
  ) %>%
  left_join(
    axial_metadata_nosurface %>%
      select(Sample, sample_type),
    by = "Sample"
  ) %>%
  group_by(Sample) %>%
  mutate(
    relative_abundance = sequences / sum(sequences)
  ) %>%
  ungroup()

whisker <- ggplot(
  major_taxa,
  aes(x = sample_type, y = relative_abundance)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7) +
  facet_wrap(~ taxa) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/whisker_plot.png", plot = whisker, width = 8, height = 10, dpi = 600, units = "in")

```

#ASV table
```{r}
ASV_table <- tax_asv_metadata %>% 
  filter(Class %in% c("Gyrista", "Bigyra", "Radiolaria")) %>% 
  filter(Supergroup == "Haptista") %>% 
  select(FeatureID, taxa, Genus, Species) %>% 
  distinct() %>% mutate(Species = case_when(
    is.na(Species) ~ "Unknown",
    TRUE ~ Species)) 

write.csv(ASV_table, file = "~/R/SAR_ASV_table.csv", row.names = FALSE)


#
CLR_df <- axial_metadata_nosurface

clr_asv <- tax_asv_metadata %>% 
  select(FeatureID, Sample, SEQUENCE_COUNT)

clr_matrix <- clr_asv %>%
  pivot_wider(
    id_cols = FeatureID,
    names_from = Sample,
    values_from = SEQUENCE_COUNT,
    values_fill = 0
  ) %>%
  column_to_rownames("FeatureID") %>%
  as.matrix()

clr_matrix <- clr(t(clr_matrix))
clr_matrix <- t(clr_matrix)

sample_meta <- CLR_df %>%
  distinct(Sample, sample_type, depth_location, sample_name)

tax_meta_definedASV <- tax_asv_metadata %>%
  filter(FeatureID %in% clr_asv$FeatureID) %>% 
  distinct(FeatureID, taxa, Species) %>% 
  mutate(Species = case_when(
      is.na(Species) ~ "Unknown",
      TRUE ~ Species)) %>% 
  mutate(ASV_Species = paste(FeatureID, Species, sep = "-"))

clr_tax_asv_definedASV <- clr_matrix %>%
  as.data.frame() %>%
  rownames_to_column("FeatureID") %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "Sample",
    values_to = "clr"
  ) %>%
  left_join(sample_meta, by = "Sample") %>%
  left_join(tax_meta_definedASV, by = "FeatureID") %>%
  group_by(taxa, ASV_Species, depth_location, sample_name) %>%
  summarise(avg_clr = mean(clr), .groups = "drop") %>% 
  arrange(taxa, ASV_Species) %>% 
  mutate(ASV_Species = factor(ASV_Species, levels = unique(ASV_Species)))

clr_tax_asv_definedASV <- clr_tax_asv_definedASV %>% 
  filter(taxa %in% c("Stramenopiles-Bigyra", "Stramenopiles-Gyrista", 
                     "Haptista", "Rhizaria-Radiolaria"))

clr_ASV_big <- clr_tax_asv_definedASV %>% 
  filter(taxa == "Stramenopiles-Bigyra")

ASV_level_big <- ggplot(clr_ASV_big, aes(x = depth_location, y = ASV_Species, fill = avg_clr)) +
  geom_tile() +
 facet_grid(
  rows = vars(taxa),
  cols = vars(sample_name),
  scales = "free",
  space = "free") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/asv_level_big.png", plot = ASV_level_big, width = 12, height = 30, dpi = 600, units = "in")

clr_ASV_gyr <- clr_tax_asv_definedASV %>% 
  filter(taxa == "Stramenopiles-Gyrista")

ASV_level_gyr <- ggplot(clr_ASV_gyr, aes(x = depth_location, y = ASV_Species, fill = avg_clr)) +
  geom_tile() +facet_grid(
  rows = vars(taxa),
  cols = vars(sample_name),
  scales = "free",
  space = "free") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/asv_level_gyr.png", plot = ASV_level_gyr, width = 12, height = 30, dpi = 600, units = "in")

clr_ASV_hap <- clr_tax_asv_definedASV %>% 
  filter(taxa == "Haptista")

ASV_level_hap <- ggplot(clr_ASV_hap, aes(x = depth_location, y = ASV_Species, fill = avg_clr)) +
  geom_tile() +
  facet_grid(
  rows = vars(taxa),
  cols = vars(sample_name),
  scales = "free",
  space = "free") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/asv_level_hap.png", plot = ASV_level_hap, width = 12, height = 30, dpi = 600, units = "in")

clr_ASV_rad <- clr_tax_asv_definedASV %>% 
  filter(taxa == "Rhizaria-Radiolaria")

ASV_level_rad <- ggplot(clr_ASV_rad, aes(x = depth_location, y = ASV_Species, fill = avg_clr)) +
  geom_tile() +
  facet_grid(
  rows = vars(taxa),
  cols = vars(sample_name),
  scales = "free",
  space = "free") +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 0) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank())

ggsave("~/R/Axial2023/Figures/asv_level_rad.png", plot = ASV_level_rad, width = 12, height = 30, dpi = 600, units = "in")

```
#Citations
```{r}
citation("tidyverse")
```
