---
title: "axial_allcode_updated"
format: html
editor: visual
---

## Characterizing Protistan Community Composition Within Hydrothermal Plumes at Axial Seamount

#1. Setup environment

```{r}
#library(readr)
#library(tidyverse)
#library(phyloseq)
#library(readxl)

#Import data
all_metadata_jan2026 <- read_csv("~/R/Axial2023/datainput/all_metadata_jan2026_v2.csv")
axial_decontam_DF <- read_excel("~/R/Axial2023/datainput/axial_decontam_DF.xlsx")
ASV <- read.delim("C:/Users/alexis98adams/Documents/R/Axial2023/datainput/clean_ASV.tsv",
                  header = TRUE, sep = "\t")
tax <- read.delim("C:/Users/alexis98adams/Documents/R/Axial2023/datainput/taxonomy.tsv", header = TRUE, sep = "\t")

axial_samplestokeep <- read_csv("~/R/Axial2023/datainput/axial_samplestokeep.csv")

axial_euk_cell_counts <- read_csv("~/R/Axial2023/datainput/axial_euk_cell_counts.csv")

#Make metadata sheet and fix naming
tax <- tax %>% rename(FeatureID = Feature.ID)

all_metadata_jan2026 <- all_metadata_jan2026 %>% 
  rename(Sample = X)

axial_metadata <- all_metadata_jan2026 %>% filter(grepl("axial", Sample,ignore.case = TRUE)) %>% filter(Sample %in% axial_samplestokeep$Sample) %>% 
  mutate(depth_category = case_when(
    (depth_category == "background") ~ "Background",
    (depth_category == "30m above") ~ "Above Plume",
    (depth_category == "under plume") ~ "Under Plume",
    (depth_category == "peak plume") ~ "Mid Plume",
    (depth_category == "plume top") ~ "Plume Top",
    (depth_category == "diffuse") ~ "Diffuse", 
    (depth_category == "euphotic") ~ "Euphotic",
    (depth_category == "oxycline") ~ "Oxycline",
    TRUE ~ depth_category))

axial_metadata <- axial_metadata %>% 
  filter(!depth_category %in% c("Oxycline", "Euphotic"))

#Keep ASVs that only have 50 or more sequences and are found in two or more samples
asv_keep <- ASV %>%
  pivot_longer(
    cols = -FeatureID,
    names_to = "sample",
    values_to = "SEQUENCE_COUNT") %>%
  group_by(FeatureID) %>%
  summarise(
    total_reads = sum(SEQUENCE_COUNT),
    prevalence = sum(SEQUENCE_COUNT > 0),
    .groups = "drop") %>%
  filter(total_reads >= 50, prevalence >= 2)

samplestokeep_ASV <- ASV %>% 
    filter(FeatureID %in% asv_keep$FeatureID) %>% 
    pivot_longer(
      cols = -c(FeatureID),
      names_to = "Sample",
      values_to = "SEQUENCE_COUNT") %>%
    filter(Sample %in% axial_metadata$Sample) 

table_asv <-  samplestokeep_ASV %>% 
  left_join(axial_metadata, by = "Sample") %>% 
  left_join(tax, by = "FeatureID") %>% 
  separate(
    col = Taxon,
    into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Genus", "Species"),
    sep = ";")
  
```

#2. Fluid Chemistry Heatmap

```{r}
FluidChemistry <- read_excel("~/R/Axial2023/datainput/2023Axial_FluidChemistry_Edited.xlsx")

FluidChemistry <- FluidChemistry[-1, ] %>% rename(Temperature = `Max T`)

###
env_var <- c("Temperature", "pH", "Cl", "SO4", "Na", "K", "Mg", "Ca")

##Vent heatmap DF
vent_site <- c("Hell", "Inferno", "Virgin", "Castle", "El Guapo", "HighDiva", "Anemone", 
               "Gollum", "LowDiva", "ASHES - 30m Above", "ASHES - Top of Plume", "ASHES - Peak Plume",
               "ASHES - Below Plume","Intl.Dist - 30m Above", "Intl.Dist - Top of Plume",
               "Intl.Dist - Peak Plume", "Intl.Dist - Below Plume", "Background")

vent_fluidchemistry <- FluidChemistry %>% 
  filter(Site %in% vent_site) %>% mutate(Site = factor(Site, levels = vent_site)) %>% 
  mutate(across(all_of(env_var), as.numeric))

`18S` <- ggplot(vent_fluidchemistry, aes(x = Site, y = "18S", fill = `18S`)) +
  geom_tile(color = "grey90") +
  scale_fill_gradient(
    low = "grey85",
    high = "darkslateblue") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text = element_text(color = "black"),
        panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "")

ggsave("~/`18S`.svg", plot = `18S`, width = 4, height = 2, units = "in")

temp <- ggplot(vent_fluidchemistry, aes(x = Site, y = "Temperature", fill = Temperature)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85",
    values = scales::rescale(c(2, 25, 50, 330)),
    limits = c(2, 330),
    oob = scales::squish,
    breaks = c(2, 50, 200, 300)) +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "Temp", fill = "T (°C)")

ggsave("~/temp.svg", plot = temp, width = 4, height = 1, units = "in")

ph <- ggplot(vent_fluidchemistry, aes(x = Site, y = "pH", fill = pH)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "pH")

ggsave("~/ph.svg", plot = ph, width = 4, height = 1, units = "in")

doc <- ggplot(vent_fluidchemistry, aes(x = Site, y = "DOC", fill = DOC)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "DOC (µM)")

ggsave("~/doc.svg", plot = doc, width = 4, height = 1, units = "in")

poc <- ggplot(vent_fluidchemistry, aes(x = Site, y = "POC", fill = POC)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "POC (µM)")

ggsave("~/poc.svg", plot = poc, width = 4, height = 1, units = "in")

#Cl
cl_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "Cl", fill = Cl)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "Cl (mM/Kg)")

ggsave("~/cl_plot.svg", plot = cl_plot, width = 4, height = 1, units = "in")

#So4
so4_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "SO4", fill = SO4)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "SO4 (mM/Kg)")

ggsave("~/so4_plot.svg", plot = so4_plot, width = 4, height = 1, units = "in")

#Na
na_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "Na", fill = Na)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "Na (mM/Kg)")

ggsave("~/na_plot.svg", plot = na_plot, width = 4, height = 1, units = "in")

#K
k_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "K", fill = K)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "K (mM/Kg)")

ggsave("~/k_plot.svg", plot = k_plot, width = 4, height = 1, units = "in")

#Mg
mg_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "Mg", fill = Mg)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "Mg (mM/Kg)")

ggsave("~/mg_plot.svg", plot = mg_plot, width = 4, height = 1, units = "in")

ca_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "Ca", fill = Ca)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "Ca (mM/Kg)")

ggsave("~/ca_plot.svg", plot = ca_plot, width = 4, height = 1, units = "in")

ca_plot <- ggplot(vent_fluidchemistry, aes(x = Site, y = "Ca", fill = Ca)) +
  geom_tile(color = "grey90") +
  scale_fill_gradientn(
    colors = c("darkseagreen1", "darkseagreen", "gold", "darkslateblue"),
    na.value = "grey85") +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(
    axis.text.x = element_blank(),
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()) +
  labs(x = "", y = "", fill = "Ca (mM/Kg)")

ggsave("~/ca_plot.svg", plot = ca_plot, width = 4, height = 1, units = "in")

#Prok
cell_count_order <- c("Anemone", "Gollum", "Diva", "ASHES-30m Above", "ASHES-Plume Top", "ASHES-Peak Plume",
                     "ASHES-Under Plume","International District-30m Above", "International District-Plume Top",
                     "International District-Peak Plume", "International District-Under Plume", "Background")

cell_counts_prok <- cell_counts_prok %>% 
  mutate(site = paste(location_name, depth_category, sep = "-")) %>% 
  mutate(site = case_when(
    str_detect(site, "Diva") ~ "Diva",
    str_detect(site, "Gollum") ~ "Gollum",
    str_detect(site, "Anemone") ~ "Anemone",
    str_detect(site, "Background") ~ "Background",
    TRUE ~ site)) %>% 
  mutate(site = factor(site, levels = cell_count_order))

prok_heatmap <- ggplot(cell_counts_prok, aes(x = site, y = "prok" , fill = `prok`)) +
  geom_tile(color = "grey90") +
  scale_fill_gradient(
    low = "grey85",
    high = "darkslateblue",
    labels = scales::label_scientific()) +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(axis.text.x = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_blank()) +
  labs(x = "", y = "Prok. Count", fill = "Cells/ml")

ggsave("~/prok_heatmap.svg", plot = prok_heatmap, width = 4, height = 2, units = "in")


#euk cell count
euk_cell_counts <- axial_euk_cell_counts %>% filter(EXP.ID != "EuphoticSubeuphotic") %>%
  filter(EXP.ID != "Oxycline") %>%
  filter(Vent_field %in% c("International District", "ASHES", "Background")) %>% 
  mutate(EXP.ID = case_when(
    EXP.ID == "30m above plume" ~ "30m Above",
    EXP.ID == "Top of plume" ~ "Plume Top", 
    EXP.ID == "Below plume" ~ "Under Plume",
    EXP.ID == "Peak plume" ~ "Peak Plume",
   TRUE ~ EXP.ID)) %>% 
  mutate(site = paste(Vent_field, EXP.ID, sep = "-")) %>% 
  mutate(site = case_when(
    str_detect(site, "Diva") ~ "Diva",
    str_detect(site, "Gollum") ~ "Gollum",
    str_detect(site, "Anemone") ~ "Anemone",
    str_detect(site, "Background") ~ "Background",
    TRUE ~ site)) %>% 
  mutate(site = factor(site, levels = cell_count_order))

euk_heatmap <- ggplot(euk_cell_counts, aes(x = site, y = "CELLS_PER_ML_total" , fill = `CELLS_PER_ML_total`)) +
  geom_tile(color = "grey90") +
  scale_fill_gradient(
    low = "grey85",
    high = "darkslateblue",
    labels = scales::label_scientific()) +
  theme_minimal() +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "in"))) +
  theme(axis.text.x = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_blank()) +
  labs(x = "", y = "Euk. Count", fill = "Cells/ml")

ggsave("~/euk_heatmap.svg", plot = euk_heatmap, width = 4, height = 2, units = "in")

```

#2.2 TOC, DOC, and POC

```{r}
SiteType_Order <- c("Background", "Diffuse", "Under Plume", "Peak Plume", "Plume Top", "30m Above")

axial_OC <- axial_metadata %>% select(DOC, POC, TOC, location_name, depth_category) %>% 
  filter(TOC > 0) %>% mutate(vent_field = case_when(
    (location_name == "Diva") ~ "International District", 
    TRUE ~ location_name)) %>% 
  mutate(depth_category = factor(depth_category, levels = SiteType_Order)) %>% 
  distinct(DOC, .keep_all = TRUE) #Biological replicates contain same numbers, removing duplicates

unique(axial_OC$vent_field)

ggplot(axial_OC, aes(x = DOC, y = depth_category, color = vent_field, shape = depth_category)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(
    "Background" = 15,
    "Diffuse" = 17, 
    "30m Above" = 16,
    "Plume Top" = 16,
    "Peak Plume" = 16,
    "Under Plume" = 16)) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod")) +
  theme_classic() + theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(shape = "none") +
  labs(x = "DOC (µM)", y = "")

ggsave("~/DOC_plot.png", plot = DOC_plot, width = 4, height = 4, dpi = 600, units = "in")

TOC_plot <- ggplot(axial_OC, aes(x = TOC, y = depth_category, color = vent_field, shape = depth_category)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(
    "Background" = 15,
    "Diffuse" = 17, 
    "30m Above" = 16,
    "Plume Top" = 16,
    "Peak Plume" = 16,
    "Under Plume" = 16)) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod")) +
  theme_classic() + theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(shape = "none") +
  labs(x = "TOC (µM)", y = "") 

ggsave("~/TOC_plot.png", plot = TOC_plot, width = 4, height = 4, dpi = 600, units = "in")

POC_plot <- ggplot(axial_OC, aes(x = POC, y = depth_category, color = vent_field, shape = depth_category)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(
    "Background" = 15,
    "Diffuse" = 17, 
    "30m Above" = 16,
    "Plume Top" = 16,
    "Peak Plume" = 16,
    "Under Plume" = 16)) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod",
    "Marker 113" = "tomato4",
    "Marker 33" = "indianred",
    "Bag City" = "coral2")) +
  theme_classic() + theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(shape = "none") +
  labs(x = "POC (µM)", y = "")

ggsave("~/POC_plot.png", plot = POC_plot, width = 4, height = 4, dpi = 600, units = "in")
```

#3. Alpha Diversity

```{r}
#Identify replicates 
unique(axial_metadata$depth_category)
  
alpha_div_metadata <- axial_metadata %>% 
  mutate(DepthCat_Location = paste(dive_ctd, filter, depth_category, location_name, depth, sep = "-"))

#DepthCat_Location will have identical names so I need to identify replicates. Just going to do this manually for simplicity. 
write.csv(alpha_div_metadata, "alpha_div_metadata.csv", row.names = FALSE)
  
  
#Rep's assigned. Criteria for reps: same dive, depth_category, location and depth
alpha_metadata_wReps <- read_csv("~/R/Axial2023/alpha_div_metadata.csv")

alpha_metadata_wReps <- alpha_metadata_wReps %>% mutate(DepthCat_Location = paste(DepthCat_Location, REP, sep = "-")) 

#Add DeptCat_Location from metadata to ASV table
alpha_ASV <- ASV %>% 
  filter(FeatureID %in% asv_keep$FeatureID) %>% 
  pivot_longer(cols = -FeatureID,
               names_to = "Sample", 
               values_to = "SEQUENCE_COUNT") %>% 
  filter(Sample %in% alpha_metadata_wReps$Sample) %>% 
  left_join(alpha_metadata_wReps %>% select(Sample, DepthCat_Location), by = "Sample")

alpha_metadata_wReps <- alpha_metadata_wReps %>% column_to_rownames("DepthCat_Location")

ASV_wide <- alpha_ASV %>%
  select(FeatureID, DepthCat_Location, SEQUENCE_COUNT) %>%
  pivot_wider(
    names_from = DepthCat_Location,
    values_from = SEQUENCE_COUNT,
    values_fill = 0)
  
#Phyloseq objects need to have row.names. Define row names from ASV column.
#Then transform into matrices. Metadata can be left as DF. 
asv_mat <- ASV_wide %>%  tibble::column_to_rownames("FeatureID") %>% as.matrix()
tax_mat <- tax %>% tibble::column_to_rownames("FeatureID") %>% as.matrix()

#Transform into phyloseq objects
ASV_phy = otu_table(asv_mat, taxa_are_rows = TRUE)
tax_phy = tax_table(tax_mat)
samples = sample_data(alpha_metadata_wReps)

#Final axial phyloseq object
Axial_phy <- phyloseq(ASV_phy, tax_phy, samples)

#Calculate alpha diversity
axial_alpha_div <- estimate_richness(Axial_phy, measures = c("Shannon", "Simpson")) %>% 
  rownames_to_column(var = "SAMPLE") %>% 
  separate(SAMPLE,
    into = c("DIVE","FILTER","DEPTH_CATEGORY", "Location_Name", "DEPTH", "REP"),  
    sep = "-", 
    remove = TRUE)

axial_alpha_div <- axial_alpha_div %>% 
    mutate(REP_ID = paste(DIVE,FILTER, Location_Name, DEPTH_CATEGORY, DEPTH, sep = "-"))  %>% 
  group_by(REP_ID) %>% 
  summarise(
    Shannon_mean = mean(Shannon, na.rm = TRUE),
    Simpson_mean = mean(Simpson, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop")

SiteType_Order <- c("Diffuse", "Under Plume", "Mid Plume", "Plume Top", "Above Plume", "Background")

axial_alpha_div <- axial_alpha_div %>% 
  separate(REP_ID,
    into = c("DIVE","FILTER", "Location_Name","DEPTH_CATEGORY", "DEPTH"),  
    sep = "-", 
    remove = TRUE)

axial_alpha_div <- axial_alpha_div %>% 
  mutate(DEPTH_CATEGORY = factor(DEPTH_CATEGORY, levels = SiteType_Order)) %>% 
  mutate(Site = case_when(
      Location_Name %in% c("Anemone", "Gollum") ~ "ASHES Vent",
      Location_Name == "Diva" ~ "International District Vent",
      TRUE ~ Location_Name)) %>% 
  mutate(Site = case_when(
    Site == "ASHES" ~ "ASHES Plume", 
    Site == "International District" ~ "International District Plume",
    TRUE ~ Site)) %>% 
  mutate(vent_field = case_when(
    str_detect(Site, "ASHES") ~ "ASHES",
    str_detect(Site, "International") ~ "International District",
    str_detect(Site, "Background") ~ "Background"))

unique(axial_alpha_div$DEPTH_CATEGORY)

#shannon plot
shannon_plot_updated <- ggplot(axial_alpha_div, aes(x = Shannon_mean, y = DEPTH_CATEGORY, color = vent_field,
    shape = Site)) +
  geom_point(aes(color = vent_field, shape = DEPTH_CATEGORY), size = 3) +
  scale_color_manual(name = "Location", values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod")) +
  scale_shape_manual(name = "Depth Category", values = c(
    "Above Plume" = 23,
    "Plume Top" = 25,
    "Mid-Plume" = 21,
    "Under Plume" = 4,
    "Diffuse" = 24,
    "Background" = 22)) +
  theme_classic() + theme(
    legend.position = "right",
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(fill = guide_legend(reverse = TRUE, ncol = 1)) +
  labs(x = "Shannon Diversity", y = "")

ggsave("~/shannon_plot_updated.png", plot = shannon_plot_updated, width = 6, height = 4, dpi = 600, units = "in")

#Simpson plot
simpson_plot_updated <- ggplot(axial_alpha_div, aes(x = Simpson_mean, y = DEPTH_CATEGORY, color = Site, shape = Site)) +
geom_point(aes(color = vent_field, shape = DEPTH_CATEGORY), size = 3) +
  scale_color_manual(name = "Location", values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod")) +
  scale_shape_manual(name = "Depth Category", values = c(
    "Above Plume" = 23,
    "Plume Top" = 25,
    "Mid-Plume" = 21,
    "Under Plume" = 4,
    "Diffuse" = 24,
    "Background" = 22)) +
  theme_classic() + theme(
    legend.position = "right",
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(fill = guide_legend(reverse = TRUE, ncol = 1)) +
  labs(x = "Simpson Diversity", y = "")

ggsave("~/simpson_plot_updated.png", plot = simpson_plot_updated, width = 6, height = 4, dpi = 600, units = "in")
```

#4a. Prokaryotic Cell Abundance plot

```{r}
cell_counts_prok <- axial_metadata %>% select(location_name, depth_category, prok) %>% 
  mutate(site = case_when(
  location_name == "Diva" ~ "International District", 
  location_name == "Anemone" ~ "ASHES",
  location_name == "Gollum" ~ "ASHES", 
  TRUE ~ location_name)) %>% mutate(depth_category = factor(depth_category, levels = SiteType_Order)) %>% filter()

prok_count_plot <- ggplot(cell_counts_prok, aes(x = prok, y = depth_category, color = site, shape = depth_category)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(
    "Background" = 15,
    "Diffuse" = 17, 
    "30m Above" = 16,
    "Plume Top" = 16,
    "Peak Plume" = 16,
    "Under Plume" = 16)) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod")) +
  theme_classic() + theme(
    legend.position = "right",
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(shape = "none") +
  scale_x_continuous(labels = scales::scientific) +
  theme(legend.title = element_blank()) +
  labs(x ="Cells/ml (log10)", y = "", title = "Prokaryotic Cell Abundance") 

ggsave("~/prok_count_plot.png", plot = prok_count_plot, width = 6, height = 4, dpi = 600, units = "in") 

```

#4b Eukaryotic Cell Count

```{r}
axial_euk_cell_counts <- read_csv("~/R/Axial2023/datainput/axial_euk_cell_counts.csv")

Cell_Site_Order <- c("Background", "Above vent", "Diffuse", "Below plume", "Peak plume",
                     "Top of plume", "30m above plume")

cell_counts <- axial_euk_cell_counts %>% filter(EXP.ID != "EuphoticSubeuphotic") %>%
  filter(EXP.ID != "Oxycline") %>% 
  mutate(Sample_type = factor(Sample_type, levels = Cell_Site_Order)) %>% 
  filter(Vent_field %in% c("ASHES", "International District", "Background"))

unique(cell_counts$Vent_field)
  
euk_count_plot <- ggplot(cell_counts, aes(x = CELLS_PER_ML_total, y = Sample_type, color = Vent_field,                             shape = Sample_type)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(
    "Background" = 15,
    "Diffuse" = 17,
    "Above vent" = 17,
    "30m above plume" = 16,
    "Top of plume" = 16,
    "Peak plume" = 16,
    "Below plume" = 16)) +
  scale_color_manual(values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "goldenrod")) +
  theme_classic() + theme(
    legend.position = "right",
    legend.text = element_text(color = "black"),
    strip.background = element_blank(),
    axis.title = element_text(color = "black", size = 9),
    axis.text.y = element_text(color = "black", vjust = 0.5, hjust = 0, size = 9),
    panel.grid.major = element_line(color = "grey70", linetype = "dashed")) +
  guides(shape = "none") +
  theme(legend.title = element_blank()) +
  labs(x ="Cells/ml", y = "", title = "Eukaryotic Cell Abundance") 

ggsave("~/euk_count_plot.png", plot = euk_count_plot, width = 6, height = 4, dpi = 600, units = "in") 



```

#5a. Upset plot

```{r}
upset_asv <- table_asv %>% 
#Refine location category for easy presence/absence categories
  mutate(vent_field = case_when(
      location_name %in% c("Anemone", "Gollum") ~ "ASHES",
      location_name == "Diva" ~ "International District",
      TRUE ~ location_name)) %>% 
  mutate(depth_category = case_when(
    depth_category == "Diffuse" ~ location_name,
    TRUE ~ depth_category)) %>% 
  mutate(full_location_name = paste(depth_category, vent_field, sep = " ")) %>% 
  mutate(full_location_name = case_when(
    full_location_name == "Background Background" ~ "Background",
    TRUE ~ full_location_name))

unique(upset_asv$full_location_name)


alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")

#Curate taxonomy
upset_asv <- upset_asv %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(Phylum != "Opisthokonta") %>% 
  mutate(Phylum = case_when( #Fix phylum
    str_detect(Phylum, "_X$") ~ paste("Unknown"),
    is.na(Phylum) ~ paste("Unknown"),
    TRUE ~ Phylum)) %>% 
  mutate(Class = case_when( #Fix class
   str_detect(Class, "_X$") ~ paste("Unknown"),
    is.na(Class) ~ paste("Unknown"),
    TRUE ~ Class)) %>% 
  mutate(supergroup_phylum = paste(Supergroup, Phylum, sep = "-")) %>% 
  mutate(phylum_class = paste(Phylum, Class, sep = "-")) %>% 
  filter(phylum_class != "Streptophyta-Unknown") %>% ##Land plants???
  mutate(taxa = case_when( #Curate taxonomy
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Class == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Class == "Radiolaria" ~ "Rhizaria-Radiolaria",
    phylum_class == "Unknown-Unknown" ~ "Unknown Eukaryota",
    str_detect(phylum_class, "Unknown-Amoebozoa_XX|Tubulinea|Evosea") ~ "Amoebozoa",
    phylum_class %in% alv ~ "Alveolata-Other",
    phylum_class == "Stramenopiles-Unknown" ~ "Stramenopiles-Other",
    str_detect(phylum_class, "Discoba|Metamonada") ~"Excavata",
    str_detect(phylum_class, "Cryptophyta|Kathablepharida") ~"Cryptista",
    str_detect(phylum_class, "Chlorophyta|Prasinodermophyta|Picozoa") ~"Archaeplastida",
    str_detect(phylum_class, "Apusomonada|Breviatea|Amoebozoa") ~"Amorphea",
    str_detect(phylum_class, "Centroplasthelida|Haptophyta") ~"Haptista",
    str_detect(phylum_class, "Nibbleridia") ~"Provora-Nibbleridia",
    str_detect(phylum_class, "Collodictyonidae|Mantamonadidea") ~ "CRuMS",
    str_detect(phylum_class, "Telonemia") ~ "Telonemia",
    TRUE ~ phylum_class))



taxa <- c("Alveolata-Dinoflagellata" = "cornflowerblue","Rhizaria-Radiolaria" = 
            "coral4","Alveolata-Ciliophora" = "coral", 
          "Stramenopiles-Gyrista" = "chocolate","Rhizaria-Cercozoa" = "cadetblue","Stramenopiles-Bigyra" = "aquamarine","Archaeplastida" = "firebrick4", 
          "Discosea-Unknown" = "plum4","Stramenopiles-Other" = "lightsalmon", "Amorphea" = "darkolivegreen3","Haptista" = "gold","Amoebozoa" = "lightgoldenrod",
          "Unknown Eukaryota" = "mistyrose","Telonemia" = "rosybrown","Alveolata-Other" = "peachpuff4","Cryptista" = "coral3","CRuMS" = "darkcyan",
          "Excavata" = "orchid","Provora-Nibbleridia" = "powderblue")

lessTax_upset_asv <- upset_asv  %>% 
  mutate(full_location_name = factor(full_location_name, levels = location_order)) %>% 
  filter(!is.na(full_location_name))

unique(alpha_metadata_wReps$DepthCat_Location)

# Step 1: Create presence table (ASV × Location)
lessTax_asv_long <- upset_asv %>%
  group_by(FeatureID, full_location_name, taxa) %>%
  summarise(present = any(SEQUENCE_COUNT > 0), .groups = "drop") %>%
  filter(present)

# Step 2: Collapse to 1 row per ASV, listing locations
lessTax_asv_upset_input <- lessTax_asv_long %>%
  group_by(FeatureID, taxa) %>%
  summarise(locations = list(unique(as.character(full_location_name))), .groups = "drop") 

library(ggupset)

# Step 3: Plot
upset_updated <- ggplot(lessTax_asv_upset_input, aes(x = locations)) +
  geom_bar(aes(fill = taxa)) +
  scale_fill_manual(values = taxa) +
  scale_x_upset(order_by = "freq", n_intersections = 25) +
  labs(x = "", y = "Number of Shared ASVs") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "Shared ASVs", fill = "Taxonomy") +
  theme_linedraw() +
  theme(axis.text.y = element_text(color="black", size=14, face = "bold"),
        axis.text.x = element_text(color="black", size=14, face = "bold"),
        axis.title = element_text(color="black", size=14, face = "bold"),
        legend.text = element_text(color = "black", size = 12),
        legend.title = element_text(color="black", size=14, face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "right")


ggsave("~/upset_updated.png", plot = upset_updated, width = 10, height = 8, dpi = 600, units = "in")

```

#5b. Venn diagrams

```{r}
#Venn
venn_asv <- upset_asv %>% 
  mutate(SAMPLE_ID = paste(vent_field, sample_type)) %>% 
  mutate(SAMPLE_ID = case_when(
    SAMPLE_ID == "Background background" ~ "Background",
    TRUE ~ SAMPLE_ID))

venn <- venn_asv %>% 
  select(FeatureID, SEQUENCE_COUNT, SAMPLE_ID) %>% 
  group_by(SAMPLE_ID) %>% 
  summarise(unique_ASV = list(unique(FeatureID))) %>% 
  deframe()

#Get ASVs for each site
ASHESplume_ASVs <- venn_asv %>% 
  filter(SAMPLE_ID == "ASHES plume", SEQUENCE_COUNT > 0) %>% 
  pull(FeatureID) %>% 
  unique()

ASHESvent_ASVs <- venn_asv %>% 
  filter(SAMPLE_ID == "ASHES vent", SEQUENCE_COUNT > 0) %>% 
  pull(FeatureID) %>% unique()

intldistPlume_ASVs <- venn_asv %>% 
  filter(SAMPLE_ID == "International District plume", SEQUENCE_COUNT > 0) %>% 
  pull(FeatureID) %>% unique()

intldistVent_ASVs <- venn_asv %>% 
  filter(SAMPLE_ID == "International District vent", SEQUENCE_COUNT > 0) %>% 
  pull(FeatureID) %>% unique()

background_ASVs <- venn_asv %>% 
  filter(SAMPLE_ID == "Background", SEQUENCE_COUNT > 0) %>% 
  pull(FeatureID) %>% unique()

shared_intldist_VentPlume <- intersect(intldistVent_ASVs, intldistPlume_ASVs)
length(shared_intldist_VentPlume)

shared_ASHES_VentPlume <- intersect(ASHESvent_ASVs, ASHESplume_ASVs)
length(shared_ASHES_VentPlume)

shared_ASHES_PlumeBackground <- intersect(ASHESplume_ASVs, background_ASVs)
length(shared_ASHES_PlumeBackground)

shared_intldist_PlumeBackground <- intersect(intldistPlume_ASVs, background_ASVs)
length(shared_intldist_PlumeBackground)

install.packages("ggvenn")
library(ggvenn)

ASHESvenn <- ggvenn(
  list(
    Plume = ASHESplume_ASVs, 
    Vent = ASHESvent_ASVs,
    Background = background_ASVs), 
  fill_color = c("orangered3", "indianred", "goldenrod")) +
  labs(title = "ASHES")

ggsave("~/ASHESvenn.png", plot = ASHESvenn, width = 10, height = 6, dpi = 600, units = "in")

IntlDist_Venn <- ggvenn(
  list(
    Plume = intldistPlume_ASVs,
    Vent = intldistVent_ASVs,
    Background = background_ASVs), 
  fill_color = c("cyan", "lightseagreen", "goldenrod")) +
  labs(title = "International District")

ggsave("~/IntlDist_Venn.png", plot = IntlDist_Venn, width = 10, height = 6, dpi = 600, units = "in")

install.packages("patchwork")
library(patchwork)

venn_patch <- IntlDist_Venn/ASHESvenn
venn_patch

ggsave("~/venn_patch.svg", plot = venn_patch, width = 20, height = 10, units = "in")

#Plume venn diagrams
```

#6. Stacked Bar Plot summarizing relative sequence abundance

```{r}
ASHES_order <- c("Gollum ", "Anemone ", "Under Plume ", "Mid Plume ", "Plume Top ", "Above Plume ", "Background")

stackedbar_axial <- upset_asv %>% #Including depth in full_location_name will ensure there are three background samples. In this case, we don't want that. 
  group_by(full_location_name, taxa) %>% #Collapse replicates
  summarise(seq_sum = sum(SEQUENCE_COUNT, na.rm = TRUE), .groups = "drop") %>% 
  group_by(full_location_name) %>% #Get total sequences count per sample
  mutate(sample_seq_total = sum(seq_sum),
         taxa_relAbun = seq_sum / sample_seq_total) %>% 
  ungroup() %>% 
  mutate(vent_field = case_when(
   str_detect(full_location_name, "ASHES") ~ "ASHES",
    str_detect(full_location_name, "International") ~ "International District",
    str_detect(full_location_name, "Background") ~ "Background")) %>% 
  mutate(full_location_name = str_remove_all(
    full_location_name,
    "ASHES|International District")) 

ASHES_stacked <- stackedbar_axial %>% 
  filter(vent_field %in% c("ASHES", "Background")) %>% 
  mutate(full_location_name = factor(full_location_name, levels = ASHES_order))


IntlDist_order <- c("Under Plume ", "Mid Plume ", "Plume Top ", "Above Plume ","Diva ")

IntlDist_stacked <- stackedbar_axial %>% 
  filter(vent_field == "International District") %>% 
  mutate(full_location_name = factor(full_location_name, levels = IntlDist_order))


#ASHES and background plot
ASHES_stacked <- ggplot(ASHES_stacked, aes(y = full_location_name, x = taxa_relAbun, fill = taxa)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = taxa) +
  theme_minimal(base_size = 12) +
  theme(
    # Text
    axis.text.x = element_text(face = "bold", color = "black", size = 12),
    axis.text.y = element_text(face = "bold", color = "black", size = 11),
    axis.title = element_text(color = "black", face = "bold"),
    strip.text = element_text(color = "black", face = "bold", size = 12),
    legend.text = element_text(color = "black", size = 14),
    
    # Backgrounds
     plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    strip.background = element_rect(fill = "gray", color = NA),
    
    # Gridlines
    panel.grid = element_blank(),
    
    # Legend
    legend.position = "right",
    legend.title = element_blank()) +
  labs(
    title = "",
    x = "",
    y = "") +
  guides(fill = guide_legend(title = NULL))

#International District
IntlDist_stacked <- ggplot(IntlDist_stacked, aes(y = full_location_name, x = taxa_relAbun, fill = taxa)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = taxa) +
  theme_minimal(base_size = 12) +
  theme(
    # Text
    axis.text.x = element_text(face = "bold", color = "black", size = 12),
    axis.text.y = element_text(face = "bold", color = "black", size = 11),
    axis.title = element_text(color = "black", face = "bold"),
    strip.text = element_text(color = "black", face = "bold", size = 12),
    legend.text = element_text(color = "black", size = 14),
    
    # Backgrounds
     plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    strip.background = element_rect(fill = "gray", color = NA),
    
    # Gridlines
    panel.grid = element_blank(),
    
    # Legend
    legend.position = "right",
    legend.title = element_blank()) +
  labs(
    title = "",
    x = "",
    y = "") +
  guides(fill = guide_legend(title = NULL))

ggsave("~/IntlDist_stacked.png", plot = IntlDist_stacked, width = 10, height = 6, dpi = 600, units = "in")

ggsave("~/ASHES_stacked.png", plot = ASHES_stacked, width = 10, height = 6, dpi = 600, units = "in")

ggsave("~/IntlDist_stacked.svg", plot = IntlDist_stacked, width = 10, height = 6, units = "in")

ggsave("~/ASHES_stacked.svg", plot = ASHES_stacked, width = 10, height = 6, units = "in")

```

#Bubble plot of total ASV count

```{r}
#one row per taxa with total ASVs
unique(table_asv$taxa)

bubble_df <- upset_asv %>%
  group_by(FeatureID, full_location_name, taxa) %>%
  summarise(present = any(SEQUENCE_COUNT > 0), .groups = "drop") %>%
  filter(present) %>%
  group_by(full_location_name, taxa) %>%
  summarise(total_ASVs = n_distinct(FeatureID), .groups = "drop") %>% 
  filter(taxa != "Unknown Eukaryota")

axial_bubble <- ggplot(bubble_df,
       aes(x = full_location_name,
           y = taxa,
           size = total_ASVs,
           fill = taxa)) +
  geom_point(shape = 21, color = "#525252", stroke = 0.8, alpha = 0.9) +
  scale_fill_manual(values = taxa) +
  guides(fill = guide_legend(
    override.aes = list(size = 6)  # increase this number
  )) +
  scale_size_continuous(range = c(3, 18)) +
  theme_linedraw() +
  labs(x = "",
       y = "",
       size = "Number of ASVs") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1,
                               color = "black", face = "bold", size = 11),
    axis.text.y = element_text(color = "black", face = "bold", size = 11),
    axis.title = element_text(color = "black", face = "bold", size = 12),
    legend.position = "right",
    legend.title = element_blank(),
  legend.text = element_text(size = 12))
  

ggsave("~/axial_bubble.png", plot = axial_bubble, width = 10, height = 10 , dpi = 600, units = "in")

```

#7. NMDS

```{r}
#NMDS

#Do not average replicates for NMDS for the first round to ensure they are clustering together closely

#library(vegan)

##Need ASV table and data sheet that contains only the samples I want to include in analysis (axial_metadata
axial_metadata <- axial_metadata %>% 
  column_to_rownames("Sample")

axial_asv <- ASV %>% 
  filter(FeatureID %in% asv_keep$FeatureID) %>% 
  column_to_rownames("FeatureID") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("Sample") %>% 
  filter(Sample %in% rownames(axial_metadata)) 

#Ensure there are no samples that have all zeros
axial_asv <- axial_asv %>%
  mutate(row_sum = rowSums(select(., where(is.integer)))) %>%
  filter(row_sum != 0) %>%
  select(-row_sum)

#Prep matrix
ASV_matrix <- axial_asv %>%
  column_to_rownames("Sample") %>% 
  select(where(is.integer)) %>%
  as.matrix()

#Compute Bray-Curtis and run NMDS
set.seed(123)
axial_nmds <- metaMDS(ASV_matrix, distance = "bray") ##Run 20 stress 0.11


#Extract NMDS scores
# Extract NMDS coordinates for samples
nmds_scores <- as.data.frame(scores(axial_nmds, display = "sites"))

# Add Sample names as a column
nmds_scores <- nmds_scores %>%
  rownames_to_column("Sample")

unique(axial_metadata$location_name)

NMDS_metadata <- axial_metadata %>% rownames_to_column("Sample") %>% 
  left_join(nmds_scores, by = "Sample") %>% 
  mutate(assoc_vent = case_when(
    location_name == "ASHES" ~ "ASHES Plume",
    location_name == "International District" ~ "International District Plume",
    location_name == "Diva" ~ "International District Vent",
    location_name == "Gollum" ~ "ASHES Vent",
    location_name == "Anemone" ~ "ASHES Vent",
    location_name == "Background" ~ "Background"))

NMDS_metadata <- NMDS_metadata %>%
  mutate(full_location_name = paste(location_name, depth_category, sep = "-")) %>% 
  mutate(full_location_name = case_when(
    str_detect(full_location_name, "Anemone|Gollum") ~ "ASHES Vent",
    str_detect(full_location_name, "Diva") ~ "International District Vent",
    str_detect(full_location_name, "Background") ~ "Background",
    TRUE ~ full_location_name)) %>% #Use this to make labels for plot
  mutate(vent_field = case_when(
    str_detect(full_location_name, "ASHES") ~ "ASHES",
    str_detect(full_location_name, "International") ~ "International District",
    str_detect(full_location_name, "Background") ~ "Background"))
  
  
unique(axial_metadata$depth_category)

NMDS_metadata <- NMDS_metadata %>%
  mutate(depth_category = factor(
    depth_category,
    levels = c(
      "Euphotic",
      "Oxycline",
      "Above Plume",
      "Plume Top",
      "Mid Plume",
      "Under Plume",
      "Diffuse",
      "Background")))

#Plot NMDS
axial_NMDS_updated <- ggplot(NMDS_metadata, aes(x = NMDS1, y = NMDS2)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(color = vent_field, shape = depth_category), size = 3) +
  scale_color_manual(name = "Location", values = c(
    "ASHES" = "firebrick2",
    "International District" = "darkcyan",
    "Background" = "darkgoldenrod")) +
  scale_shape_manual(name = "Depth Category", values = c(
    "Euphotic" = 8,
    "Oxycline" = 3,
    "Above Plume" = 23,
    "Plume Top" = 25,
    "Mid Plume" = 21,
    "Under Plume" = 4,
    "Diffuse" = 24,
    "Background" = 22)) +
  theme(
    axis.text.y = element_text(colour = "black", size = 10),
    axis.text.x = element_text(colour = "black", size = 10), 
    axis.title.y = element_text(face = "bold", size = 11), 
    axis.title.x = element_text(face = "bold", size = 11, colour = "black"), 
    panel.background = element_blank(), 
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
    legend.title = element_text(size = 12, colour = "black", face = "bold"),
    legend.text = element_text(size = 10, colour ="black"), 
    legend.position = "right",
    legend.key = element_blank(),
    legend.key.size = unit(0.5, "lines"),
    legend.spacing.y = unit(0.1, "lines"),
    legend.spacing.x = unit(0.1, "lines")) +
  labs(x = "NMDS1", y = "NMDS2") #+
geom_text(aes(label = full_location_name)) 

ggsave("~/axial_NMDS_updated.png", plot = axial_NMDS_updated, width = 8, height = 4, dpi = 600, units = "in")

ggsave("~/axial_NMDS_updated.svg", plot = axial_NMDS_updated, width = 8, height = 4, units = "in")
  
```

#8.1 CLR heatmap with dendogram

```{r}
#Keep ASVs that only have 50 or more sequences and are found in two or more samples
#library(compositions)

clr_asv <- ASV %>% 
  pivot_longer(
    cols = -c(FeatureID),
    names_to = "sample",
    values_to = "SEQUENCE_COUNT") %>%
  filter(FeatureID %in% asv_keep$FeatureID) %>% 
  filter(sample %in% axial_metadata$Sample) 

clr_wide <- clr_asv %>% 
  pivot_wider(id_cols = FeatureID,
    names_from = sample,
    values_from = SEQUENCE_COUNT)

clr_wide <- clr_wide %>% 
  column_to_rownames("FeatureID") 

clr_matrix <- as.matrix(clr_wide)

clr_matrix <- clr(clr_matrix)

clr_DF <- as.data.frame(clr_matrix) %>% 
  rownames_to_column("FeatureID") %>% 
  pivot_longer(cols = -FeatureID,
               names_to = "Sample",
               values_to = "clr") %>% 
  left_join(tax, by = "FeatureID") %>% select(-Consensus) %>% 
  separate(
    col = Taxon,
    into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Genus", "Species"),
    sep = ";") %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(Phylum != "Opisthokonta") %>% 
  mutate(Phylum = case_when( #Fix phylum
    str_detect(Phylum, "_X$") ~ paste("Unknown"),
    is.na(Phylum) ~ paste("Unknown"),
    TRUE ~ Phylum)) %>% 
  mutate(Class = case_when( #Fix class
   str_detect(Class, "_X$") ~ paste("Unknown"),
    is.na(Class) ~ paste("Unknown"),
    TRUE ~ Class)) %>%
  rowwise() %>%
  mutate(
    taxonomy = paste(
      na.omit(c(
        if (!is.na(Phylum) & !str_detect(Phylum, "_X+$") & Phylum != "Unknown") Phylum,
        if (!is.na(Class) & !str_detect(Class, "_X+$") & Class != "Unknown") Class,
        if (!is.na(Genus) & !str_detect(Genus, "_X+$") & Genus != "Unknown") Genus)),
      collapse = ";")) %>%
  ungroup() #I think metadata left_join accidentally got removed

clr_DF <- clr_DF %>% 
  mutate(depth_category = case_when(
    depth_category == "Diffuse" ~ location_name,
    location_name %in% c("ASHES", "International District") ~ 
    paste(location_name, depth_category),
    TRUE ~ depth_category)) 


#plotting genus not species (FeatureID) level, average all ASVs with the same taxonomy label. 

tax_matrix <- clr_DF %>%
  group_by(taxonomy, Sample) %>% 
  summarise(clr = mean(clr), .groups = "drop") %>%
  pivot_wider(names_from = Sample, values_from = clr) %>%
  column_to_rownames("taxonomy") %>%
  as.matrix() 

#Are there multiple samples for each site depth category (i.e ASHES peak plume)? If so, these CLR values should be averaged too. 

tax_matrix_depth <- clr_DF %>% 
  group_by(depth_category, taxonomy) %>% 
  summarise(clr = mean(clr), .groups = "drop") %>%
  pivot_wider(names_from = depth_category, values_from = clr) %>%
  column_to_rownames("taxonomy") %>%
  as.matrix()

tax_matrix_depth[is.na(tax_matrix_depth)] <- 0


# Rows (taxa)
##NOTE: euclidean is the distance between two points. In this case, how different the CLR values. 
#Need to calculate how different two taxa are based on their CLR profiles across plume depths
dist_rows <- dist(tax_matrix_depth, method = "euclidean")
hc_rows <- hclust(dist_rows, method = "average")

# Columns (depth categories)
dist_cols <- dist(t(tax_matrix_depth), method = "euclidean")
hc_cols <- hclust(dist_cols, method = "average")

#library(pheatmap)

pheatmap(
  tax_matrix_depth,
  cluster_rows = hc_rows,
  cluster_cols = hc_cols,
  color = colorRampPalette(c("navy", "white", "red"))(100),
  border_color = "black",
  fontsize_row = 8,
  fontsize_col = 8,
  angle_col = 45)

```

#8.2 Plume CLR heatmap

```{r}
clr_asv <- alpha_ASV %>% 
  select(-Sample) %>% 
  separate(DepthCat_Location,
           into = c("DIVE","FILTER","DEPTH_CATEGORY", "Location_Name", "DEPTH", "REP"),  
           sep = "-", 
           remove = TRUE) %>%
  filter(str_detect(DEPTH_CATEGORY, regex("plume", ignore_case = TRUE))) %>% #Regex ignores case sensitivies
  mutate(DepthCat_Location = paste(DIVE,FILTER,DEPTH_CATEGORY, Location_Name, DEPTH, REP, sep = "-")) %>% 
  select(-DIVE, -FILTER, -DEPTH_CATEGORY, -Location_Name, -DEPTH, -REP)

clr_wide <- clr_asv %>% 
  pivot_wider(id_cols = FeatureID,
              names_from = DepthCat_Location,
              values_from = SEQUENCE_COUNT)


clr_wide <- clr_wide %>% 
  column_to_rownames("FeatureID") 

clr_matrix <- as.matrix(clr_wide)

clr_matrix <- clr(clr_matrix)

clr_DF <- as.data.frame(clr_matrix) %>% 
  rownames_to_column("FeatureID") %>% 
  pivot_longer(cols = -FeatureID,
               names_to = "Sample",
               values_to = "clr") %>% 
  left_join(tax, by = "FeatureID") %>% select(-Consensus) %>% 
  separate(
    col = Taxon,
    into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Genus", "Species"),
    sep = ";") %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(Phylum != "Opisthokonta") %>% 
  mutate(Phylum = case_when( #Fix phylum
    str_detect(Phylum, "_X$") ~ paste("Unknown"),
    is.na(Phylum) ~ paste("Unknown"),
    TRUE ~ Phylum)) %>% 
  mutate(Class = case_when( #Fix class
   str_detect(Class, "_X$") ~ paste("Unknown"),
    is.na(Class) ~ paste("Unknown"),
    TRUE ~ Class)) %>%
  rowwise() %>%
  mutate(
    taxonomy = paste(
      na.omit(c(
        if (!is.na(Phylum) & !str_detect(Phylum, "_X+$") & Phylum != "Unknown") Phylum,
        if (!is.na(Class) & !str_detect(Class, "_X+$") & Class != "Unknown") Class,
        if (!is.na(Genus) & !str_detect(Genus, "_X+$") & Genus != "Unknown") Genus)),
      collapse = ";")) %>%
  ungroup() %>% separate(Sample,
           into = c("DIVE","FILTER","DEPTH_CATEGORY", "Location_Name", "DEPTH", "REP"),  
           sep = "-", 
           remove = TRUE) %>% 
  mutate(Sample = paste(DEPTH_CATEGORY, Location_Name, sep = "-")) #remove repID so it's same sample ID

tax_matrix <- clr_DF %>%
  group_by(taxonomy, Sample) %>% 
  summarise(clr = mean(clr), .groups = "drop") %>% 
  pivot_wider(names_from = Sample, values_from = clr) %>%
  column_to_rownames("taxonomy") %>%
  as.matrix() 

tax_matrix[is.na(tax_matrix)] <- 0


# Rows (taxa)
##NOTE: euclidean is the distance between two points. In this case, how different the CLR values. 
#Need to calculate how different two taxa are based on their CLR profiles across plume depths
dist_rows <- dist(tax_matrix, method = "euclidean")
hc_rows <- hclust(dist_rows, method = "average")

# Columns (depth categories)
dist_cols <- dist(t(tax_matrix), method = "euclidean")
hc_cols <- hclust(dist_cols, method = "average")

#library(pheatmap)

clr_dendo_plume <- pheatmap(
  tax_matrix,
  cluster_rows = hc_rows,
  cluster_cols = hc_cols,
  color = colorRampPalette(c("navy", "white", "red"))(100),
  border_color = "black",
  fontsize_row = 8,
  fontsize_col = 8,
  angle_col = 45)


ggsave("~/clr_dendo_plume.png", plot = clr_dendo_plume, width = 6, height = 8, dpi = 600, units = "in")


##Remove the least variable taxa to clean up the figure
install.packages("matrixStats")
library(matrixStats)

# Calculate variance across samples for each taxonomy row
taxa_variance <- rowVars(tax_matrix)

# Keep top 50 most variable taxonomies
top_taxa <- order(taxa_variance, decreasing = TRUE)[1:50]
tax_filtered <- tax_matrix[top_taxa, ]

# Recalculate clustering on filtered matrix
dist_rows_filtered <- dist(tax_filtered, method = "euclidean")
hc_rows_filtered <- hclust(dist_rows_filtered, method = "average")

dist_cols_filtered <- dist(t(tax_filtered), method = "euclidean")
hc_cols_filtered <- hclust(dist_cols_filtered, method = "average")

# Plot filtered heatmap
clr_dendo_plume_filtered <- pheatmap(
  tax_filtered,
  cluster_rows = hc_rows_filtered,
  cluster_cols = hc_cols_filtered,
  color = colorRampPalette(c("navy", "white", "red"))(100),
  border_color = "black",
  fontsize_row = 8,
  fontsize_col = 8,
  angle_col = 90
)

# Save figure
ggsave(
  "~/clr_dendo_plume_filtered.png",
  plot = clr_dendo_plume_filtered,
  width = 7,
  height = 8,
  dpi = 600,
  units = "in"
)
```

#9 Bray Curtis Dissimilarity of plume samples vs. background

```{r}
bc_asv <- alpha_ASV %>% 
  select(-Sample) %>% 
  separate(DepthCat_Location,
           into = c("DIVE","FILTER","DEPTH_CATEGORY", "Location_Name", "DEPTH", "REP"),  
           sep = "-", 
           remove = TRUE) %>%
  filter(str_detect(DEPTH_CATEGORY, regex("plume|Background", ignore_case = TRUE))) %>% #Regex ignores case sensitivies
  mutate(Sample = paste(DIVE,FILTER,DEPTH_CATEGORY, Location_Name, DEPTH, sep = "-")) %>% 
  select(-DIVE, -FILTER, -DEPTH_CATEGORY, -Location_Name, -DEPTH, -REP) %>% 
  group_by(FeatureID, Sample) %>% 
  summarise(total_seq = sum(SEQUENCE_COUNT)) 

bc_wide <- bc_asv %>% 
  pivot_wider(id_cols = FeatureID,
              names_from = Sample,
              values_from = total_seq)


bc_wide <- bc_wide %>% 
  column_to_rownames("FeatureID") 


# Raw abundance
bc_matrix <- as.matrix(bc_wide)

# Bray-Curtis
bray_dist <- vegdist(t(bc_matrix), method = "bray") #need transpose to compare samples to samples

bc_scores <- as.data.frame(as.matrix(bray_dist)) %>%
  rownames_to_column("Sample1") %>% 
  pivot_longer(cols = -Sample1, 
               names_to = "Sample2",
               values_to = "BrayCurtis") %>% 
  filter(str_detect(Sample2, regex("background", ignore_case = TRUE))) %>% #Summarize depth_category vs background
  mutate(Sample1_site = case_when(
    str_detect(Sample1, "Above Plume-ASHES") ~ "Above Plume-ASHES",
    str_detect(Sample1, "Mid Plume-ASHES") ~ "Mid Plume-ASHES",
    str_detect(Sample1, "Plume Top-ASHES") ~ "Plume Top-ASHES",
    str_detect(Sample1, "Background") ~ "Background",
    str_detect(Sample1, "Above Plume-International District") ~ "Above Plume-International District",
    str_detect(Sample1, "Mid Plume-International District") ~ "Mid Plume-International District",
    str_detect(Sample1, "Plume Top-International District") ~ "Plume Top-International District",
    str_detect(Sample1, "Under Plume-International District") ~ "Under Plume-International District",
    str_detect(Sample1, "Under Plume-ASHES") ~ "Under Plume-ASHES",
    TRUE ~ Sample1)) %>% 
  filter(Sample1_site != "Background") %>% 
  mutate(Sample2_Site = case_when(
    str_detect(Sample2, "Background") ~ "Background",
    TRUE ~ Sample2)) %>% 
    group_by(Sample1_site, Sample2_Site) %>% 
  summarise(meanBray = mean(BrayCurtis))

order <- c("Under Plume-International District", "Mid Plume-International District", "Plume Top-International District", "Above Plume-International District", "Under Plume-ASHES", "Mid Plume-ASHES", "Plume Top-ASHES", "Above Plume-ASHES")

bc_scores <- bc_scores %>% 
  mutate(Sample1_site = factor(Sample1_site, levels = order))

ASHES_bc <- bc_scores %>% 
  filter(str_detect(Sample1_site, "ASHES"))

International_bc <- bc_scores %>% 
  filter(str_detect(Sample1_site, "International"))


unique(bc_scores$Sample1_site)

ASHES_bc <- ggplot(ASHES_bc, aes(y = Sample1_site, x = meanBray)) +
  geom_point(size = 3) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Plume Depth Category",
    x = "Mean Bray-Curtis distance to Background"
  )

IntlDist_bc <- ggplot(International_bc, aes(y = Sample1_site, x = meanBray)) +
  geom_point(size = 3) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Plume Depth Category",
    x = "Mean Bray-Curtis distance to Background"
  )

bc_patch <- ASHES_bc/IntlDist_bc

ggsave(
  "~/bc_patch.png",
  plot = bc_patch,
  width = 7,
  height = 8,
  dpi = 600,
  units = "in")  
```
